---
title: 'MAT-427: KNN + Preprocessing'
author: Eric Friedlander
footer: "[🔗 MAT 427 - Spring 2025 -  Schedule](https://mat427sp25.netlify.app/schedule)"
logo: "../images/logo.png"
format: 
  revealjs:
    theme: slides.scss
    multiplex: false
    transition: fade
    slide-number: false
    incremental: false 
    chalkboard: true
execute:
  freeze: auto
  echo: true
knitr:
  opts_chunk: 
    R.options:      
    width: 200
---


## Computational Setup

```{r setup}
library(tidyverse)
library(tidymodels)
library(knitr)
library(modeldata) # contains ames dataset

tidymodels_prefer()
```


## Comparing Models: Data Splitting with `tidymodels` {.smaller}

```{r}
set.seed(427) # Why?

ames_split <- initial_split(ames, prop = 0.70, strata = Sale_Price) # initialize 70/30
ames_split

ames_train <- training(ames_split) # get training data
ames_test <- testing(ames_split) # get test data
```



## [K-Nearest Neighbors Regression (multiple predictors)]{.r-fit-text} {.smaller}

```{r}
#| echo: TRUE

ames |>
  select(Sale_Price, Gr_Liv_Area, Bedroom_AbvGr) |>
  head() |> 
  kable()
```
:::{.incremental}
- Should 1 square foot count the same as 1 bedroom?
- Need to **center and scale** (freq. just say scale)
  + subtract mean from each predictor
  + divide by standard deviation of each predictor
  + compares apples-to-apples
:::

# Why center and scale

## New observation

```{r}
#| code-fold: true
new_house <- tibble(Gr_Liv_Area = 2000, Bedroom_AbvGr = 2)
ggplot(ames_train, aes(x = Gr_Liv_Area, y = Bedroom_AbvGr, color = Sale_Price)) + 
  geom_point() +
  geom_point(data = new_house, color = "red")
```

-   Where do you think the 10 nearest neighbors *should* be?

## Computing Distances

```{r}
ames_dist <- ames_train |> 
  mutate(dist = sqrt((Gr_Liv_Area - 2000)^2 + (Bedroom_AbvGr-2)^2)) |> 
  arrange(dist)
```

## 10 "Nearest Neighbors"

```{r}
#| code-fold: true

ggplot(ames_train, aes(x = Gr_Liv_Area, y = Bedroom_AbvGr, color = Sale_Price)) + 
  geom_point() +
  geom_point(data = new_house, color = "red") +
  geom_point(data = ames_dist |> slice(1:10), color = "green")
```

-   Are they where you thought they *should* be?

## Looking at distances {.smaller}

```{r}
ames_dist |>
  select(Gr_Liv_Area, Bedroom_AbvGr, dist) |> 
  kable()
```

# Visualizing Centering and Scaling in 1-D

## Original data

```{r}
ggplot(ames_train, aes(x = Gr_Liv_Area)) +
  geom_histogram() +
  theme(text = element_text(size = 20)) 
```

## Centering

- What's different between these two plots

::::{.columns}

:::{.column}
Original data
```{r}
#| code-fold: true
ames_train |> 
ggplot(aes(x = Gr_Liv_Area)) +
  geom_histogram() +
  theme(text = element_text(size = 20)) 
```
:::

::: {.column}
Centered data
```{r}
#| code-fold: true
ames_train |> mutate(Gr_Liv_Area_Centered = Gr_Liv_Area - mean(Gr_Liv_Area)) |> 
ggplot(aes(x = Gr_Liv_Area_Centered)) +
  geom_histogram() +
  theme(text = element_text(size = 20))
```
:::
::::

## Scaling

- What's different between these two plots

::::{.columns}
:::{.column}
Original data
```{r}
#| code-fold: true
ames_train |> 
ggplot(aes(x = Gr_Liv_Area)) +
  geom_histogram() +
  theme(text = element_text(size = 20)) 
```
:::

::: {.column}
Normalized (centered and scaled) data
```{r}
#| code-fold: true
ames_train |> mutate(Gr_Liv_Area_Centered_Scaled = (Gr_Liv_Area - mean(Gr_Liv_Area))/sd(Gr_Liv_Area)) |> 
ggplot(aes(x = Gr_Liv_Area_Centered_Scaled)) +
  geom_histogram() +
  theme(text = element_text(size = 20))
```
:::
::::

# Visualizing Centering and Scaling: 2D

## Original Data

```{r}
#| code-fold: true
ggplot(ames_train, aes(x = Gr_Liv_Area, y = Bedroom_AbvGr)) + 
  geom_point()
```

## Centering

- What's different between these two plots

::::{.columns}
:::{.column}
Original Data
```{r}
#| code-fold: true
ames_train |> 
ggplot(aes(x = Gr_Liv_Area, y = Bedroom_AbvGr)) + 
  geom_point()+
  theme(text = element_text(size = 20))
```
:::

::: {.column}
Centered data
```{r}
#| code-fold: true
ames_train |> mutate(Gr_Liv_Area_Centered = Gr_Liv_Area - mean(Gr_Liv_Area),
                     Bedroom_AbvGr_Centered = Bedroom_AbvGr - mean(Bedroom_AbvGr)) |> 
ggplot(aes(x = Gr_Liv_Area_Centered, y = Bedroom_AbvGr_Centered)) +
  geom_point() +
  theme(text = element_text(size = 20))
```
:::
::::

## Scaling

- What's different between these two plots

::::{.columns}
:::{.column}
Original data
```{r}
#| code-fold: true
ames_train |> 
ggplot(aes(x = Gr_Liv_Area, y = Bedroom_AbvGr)) + 
  geom_point()+
  theme(text = element_text(size = 20))
```
:::

::: {.column}
Normalized data
```{r}
#| code-fold: true
ames_train |> mutate(Gr_Liv_Area_Scaled = scale(Gr_Liv_Area),
                     Bedroom_AbvGr_Scaled = scale(Bedroom_AbvGr)) |> 
ggplot(aes(x = Gr_Liv_Area_Scaled, y = Bedroom_AbvGr_Scaled)) +
  geom_point() +
  theme(text = element_text(size = 20))
```
:::
::::

# Dealing with new data

## New Observation

```{r}
#| code-fold: true
new_house <- tibble(Gr_Liv_Area = 2000, Bedroom_AbvGr = 2)
ggplot(ames_train, aes(x = Gr_Liv_Area, y = Bedroom_AbvGr, color = Sale_Price)) + 
  geom_point() +
  geom_point(data = new_house, color = "red")
```

-   Where should we put this point on the normalized plot?

## Plotting new observation witout normalizing

```{r}
#| code-fold: true
new_house <- tibble(Gr_Liv_Area = 2000, Bedroom_AbvGr = 2)
ames_train |> mutate(Gr_Liv_Area_Scaled = scale(Gr_Liv_Area), Bedroom_AbvGr_Scaled = scale(Bedroom_AbvGr)) |> 
  ggplot(aes(x = Gr_Liv_Area_Scaled, y = Bedroom_AbvGr_Scaled)) +
  geom_point() +
  geom_point(data = new_house, mapping = aes(Gr_Liv_Area, y = Bedroom_AbvGr), color = "red")
```

-   Does this look right?

## Normalized using training data

::::{.columns}
:::{.column}
Original Data
```{r}
#| code-fold: true
new_house <- tibble(Gr_Liv_Area = 2000, Bedroom_AbvGr = 2)
ggplot(ames_train, aes(x = Gr_Liv_Area, y = Bedroom_AbvGr)) + 
  geom_point() +
  geom_point(data = new_house, color = "red") +
  theme(text = element_text(size = 20))
```
:::

:::{.column}
Scaled data
```{r}
#| code-fold: true
ames_train_scaled <- 
  ames_train |> mutate(Gr_Liv_Area_Scaled = scale(Gr_Liv_Area), Bedroom_AbvGr_Scaled = scale(Bedroom_AbvGr))

new_house <- tibble(Gr_Liv_Area = 2000, Bedroom_AbvGr = 2) |> 
  mutate(Gr_Liv_Area_Scaled = (Gr_Liv_Area - mean(ames_train$Gr_Liv_Area))/sd(ames_train$Gr_Liv_Area), 
         Bedroom_AbvGr_Scaled = (Bedroom_AbvGr - mean(ames_train$Bedroom_AbvGr))/ sd(ames_train$Bedroom_AbvGr))

ames_train_scaled |> 
  ggplot(aes(x = Gr_Liv_Area_Scaled, y = Bedroom_AbvGr_Scaled)) +
  geom_point() +
  geom_point(data = new_house, color = "red")+
  theme(text = element_text(size = 20))
```
:::
::::

## Compute new distances

```{r}
new_house |> kable()
ames_dist <- ames_train_scaled |> 
  mutate(orig_dist = sqrt((Gr_Liv_Area - 2000)^2 + (Bedroom_AbvGr-2)^2),
         scaled_dist = sqrt((Gr_Liv_Area_Scaled - 0.99)^2 + (Bedroom_AbvGr_Scaled-(-1.04))^2))

orig_closest <- ames_dist |> 
  arrange(orig_dist) |> 
  head(10)

scaled_closest <- ames_dist |> 
  arrange(scaled_dist) |> 
  head(10)
```

## Scaled distances look better {.smaller}

:::: columns
::: column
```{r}
orig_closest |> select(Gr_Liv_Area, Bedroom_AbvGr, Gr_Liv_Area_Scaled, Bedroom_AbvGr_Scaled, orig_dist, scaled_dist) |> kable()
```
:::

::: column
```{r}
scaled_closest |>  select(Gr_Liv_Area, Bedroom_AbvGr, Gr_Liv_Area_Scaled, Bedroom_AbvGr_Scaled, orig_dist, scaled_dist) |>  kable()
```
:::
::::

## Which plot looks better {.smaller}

:::: columns
::: column
Original distances
```{r}
#| code-fold: true
ggplot(ames_train, aes(x = Gr_Liv_Area, y = Bedroom_AbvGr, color = Sale_Price)) + 
  geom_point() +
  geom_point(data = new_house, color = "red") +
  geom_point(data = orig_closest, color = "green") +
  theme(text = element_text(size = 20))
```
:::

::: column
Scaled distances
```{r}
#| code-fold: true
ggplot(ames_train, aes(x = Gr_Liv_Area, y = Bedroom_AbvGr, color = Sale_Price)) + 
  geom_point() +
  geom_point(data = new_house, color = "red") +
  geom_point(data = scaled_closest, color = "green") +
  theme(text = element_text(size = 20))
```
:::
::::

# Dealing with the test set


## Scaling Test Set

-   Subtract mean of corresponding variable in *training* set
-   Divide by sd of corresponding variable in *training* set

## Visualizing Training and Test Sets: Untransformed

```{r}
#| code-fold: true
ggplot(ames_train, aes(x = Gr_Liv_Area, y = Bedroom_AbvGr)) + 
  geom_point(color = "red", alpha = 0.2) +
  geom_point(data = ames_test, color = "blue", alpha = 0.2) +
  theme(text = element_text(size = 20))
```

## Visualizing Training and Test Sets: Self Centered

:::: columns
::: column
Original Data
```{r}
#| code-fold: true
ggplot(ames_train, aes(x = Gr_Liv_Area, y = Bedroom_AbvGr)) + 
  geom_point(color = "red", alpha = 0.2) +
  geom_point(data = ames_test, color = "blue", alpha = 0.2) +
  theme(text = element_text(size = 20))
```
:::

::: column
Scaled Data
```{r}
#| code-fold: true
ames_train_centered <- ames_train |> 
  mutate(Gr_Liv_Area_Scaled = scale(Gr_Liv_Area),
         Bedroom_AbvGr_Scaled = scale(Bedroom_AbvGr))
ames_test_self_centered <- ames_test |> 
  mutate(Gr_Liv_Area_Scaled = scale(Gr_Liv_Area),
         Bedroom_AbvGr_Scaled = scale(Bedroom_AbvGr))
ggplot(ames_train_centered, aes(x = Gr_Liv_Area_Scaled, y = Bedroom_AbvGr_Scaled)) + 
  geom_point(color = "red", alpha = 0.2) +
  geom_point(data = ames_test_self_centered, color = "blue", alpha = 0.2) +
  theme(text = element_text(size = 20))
```
:::
::::

## Visualizing Training and Test Sets: Training Centered

:::: columns
::: column
Original data
```{r}
#| code-fold: true
ggplot(ames_train, aes(x = Gr_Liv_Area, y = Bedroom_AbvGr)) + 
  geom_point(color = "red", alpha = 0.2) +
  geom_point(data = ames_test, color = "blue", alpha = 0.2) +
  theme(text = element_text(size = 20))
```
:::

::: column
Scaled data
```{r}
#| code-fold: true
ames_train_centered <- ames_train |> 
  mutate(Gr_Liv_Area_Scaled = scale(Gr_Liv_Area),
         Bedroom_AbvGr_Scaled = scale(Bedroom_AbvGr))
ames_test_self_centered <- ames_test |> 
  mutate(Gr_Liv_Area_Scaled = (Gr_Liv_Area - mean(ames_train$Gr_Liv_Area))/sd(ames_train$Gr_Liv_Area), 
         Bedroom_AbvGr_Scaled = (Bedroom_AbvGr - mean(ames_train$Bedroom_AbvGr))/ sd(ames_train$Bedroom_AbvGr))
ggplot(ames_train_centered, aes(x = Gr_Liv_Area_Scaled, y = Bedroom_AbvGr_Scaled)) + 
  geom_point(color = "red", alpha = 0.2) +
  geom_point(data = ames_test_self_centered, color = "blue", alpha = 0.2) +
  theme(text = element_text(size = 20))
```
:::
::::

# Recipes and Workflows

## Preprocessing + Feature Engineering {.smaller}

-   Preprocessing: reformatting and transforming data so it can be used for modeling
    +   Centering and scaling
    +   Converting categorical data into dummy variables or the correct format
    +   Dealing with missing values
-   Feature Engineering: reformatting and transforming data to improve the performance of your model
    +   Combining variables and dimension reduction techniques
    +   Transforming variables (e.g. squaring or logging)
-   For recommended preprocessing steps see [Appendix A of TMWR](https://www.tmwr.org/pre-proc-table)
-   `recipes`: combine feature engineering and preprocessing steps into single object that can be applied to different data sets

## Model workflow

-   Start with raw data
-   Preprocessing and feature engineering steps
-   Fit "model"
-   Make predictions using model

## Where does my model start?

-   Start with raw data
-   Model starts here
    -   Preprocessing and feature engineering steps
    -   Fit "model"
-   Make predictions using model

## Model Assessment

-   Start with raw data
-   Split Data
-   Using training data
    -   Preprocessing and feature engineering steps
    -   Fit "model"
-   Make predictions on test set
-   Compute error metrics

## `workflows` and `recipes` in `tidymodels`

-   `tidymodels` provides `workflows` that combine model fitting with preprocessing steps and consist of:
    -   A model object
    -   A `recipe` which combines all of the preprocessing steps

## Creating a KNN model 

```{r}
knn10_model <- nearest_neighbor(neighbors = 10) |>   # 10-nn regression
  set_engine("kknn") |>
  set_mode("regression")
```

## `recipe`'s in `tidymodels`

1.  Call `receipe` with R formula and training set to assign features roles (predictor vs. outcome).
2.  Sequence of `step_*` specifying the list of transformations.
3.  Apply `recipe` to new data using `predict`.

## Center and Scale Recipe

```{r}
ames_preproc <- recipe(Sale_Price ~ Bedroom_AbvGr + Gr_Liv_Area, data = ames_train) |> 
  step_normalize(all_numeric_predictors())
```

## Create `workflow`

```{r}
knn10_workflow <- workflow() |> 
  add_model(knn10_model) |> 
  add_recipe(ames_preproc)
```

## Fitting Model


```{r}
knnfit10 <- knn10_workflow |> 
  fit(data = ames_train)
```

## Making predictions: Single observation

Test Point: `Gr_Liv_area` = 2000 square feet, and `Bedroom_AbvGr` = 3, then

```{r}
new_house |> kable()

# obtain 10-nn prediction
predict(knnfit10, new_data = new_house) |> kable()
```

## Making Predictions: Test Set

```{r}
# obtain 10-nn prediction
predict(knnfit10, new_data = ames_test) |> head() |> kable()
```


## [Linear Regression vs K-Nearest Neighbors]{.r-fit-text} {.smaller}

- Linear regression is a parametric approach (with restrictive assumptions), KNN is non-parametric.
- Linear regression works for regression problems ($Y$ numerical), KNN can be used for both regression and classification - i.e. $Y$ qualitative
- Linear regression is interpretable, KNN is not.
- Linear regression can accommodate qualitative predictors and can be extended to include interaction terms as well while KNN does not allow for qualitative predictors
- Performance: KNN can be pretty good for small $p$, that is, $p \le 4$ and large $n$. Performance of KNN deteriorates as $p$ increases - *curse of dimensionality*

# Choosing a model

## Linear regression

```{r}
# Best model from last time
fit3 <- linear_reg() |> 
  set_engine("lm") |> 
  fit(Sale_Price ~ Gr_Liv_Area + Bedroom_AbvGr, data = ames_train)
```

## 5-NN

```{r}
knn5_model <- nearest_neighbor(neighbors = 5) |> 
  set_engine("kknn") |> 
  set_mode("regression")

knn5_fit <- workflow() |> 
  add_model(knn5_model) |> 
  add_recipe(ames_preproc) |> 
  fit(data = ames_train)
```

## Making test predictions

```{r}
ames_test_preds <- ames_test |> 
  mutate(lr_preds = predict(fit3, new_data = ames_test)$.pred,
         knn5_preds = predict(knn5_fit, new_data = ames_test)$.pred,
         knn10_preds = predict(knnfit10, new_data = ames_test)$.pred)
```

## Computing Error Metrics: RMSE

```{r}
rmse(ames_test_preds, estimate = lr_preds, truth = Sale_Price) |> kable()
rmse(ames_test_preds, estimate = knn5_preds, truth = Sale_Price) |> kable()
rmse(ames_test_preds, estimate = knn10_preds, truth = Sale_Price) |> kable()
```

## Computing Error Metrics: $R^2$

```{r}
rsq(ames_test_preds, estimate = lr_preds, truth = Sale_Price) |> kable()
rsq(ames_test_preds, estimate = knn5_preds, truth = Sale_Price) |> kable()
rsq(ames_test_preds, estimate = knn10_preds, truth = Sale_Price) |> kable()
```

