---
title: 'MATH 427: Classification + Logistic Regression'
author: Eric Friedlander
footer: "[🔗 MAT 427 - Spring 2025 -  Schedule](https://mat427sp25.netlify.app/schedule)"
logo: "../images/logo.png"
format: 
  revealjs:
    theme: slides.scss
    multiplex: false
    transition: fade
    slide-number: false
    incremental: false 
    chalkboard: true
execute:
  freeze: auto
  echo: true
knitr:
  opts_chunk: 
    R.options:      
    width: 200
---

```{r}
#| include: FALSE
library(tidyverse)
library(tidymodels)
library(gridExtra)
library(janitor) # for next contingency tables
library(kableExtra)
library(ISLR2)

tidymodels_prefer()
```

## Classification Problems {.smaller}

- Response $Y$ is qualitative (categorical).
- Objective: build a classifier $\hat{Y}=\hat{C}(\mathbf{X})$
  + assigns class label to a future unlabeled (unseen) observations
  + understand the relationship between the predictors and response
- Two ways to make predictions
  + Class probabilities
  + Class labels

## Default Dataset {.smaller}


A simulated data set containing information on ten thousand customers. The aim here is to predict which customers will default on their credit card debt.

```{r}
#| message: false

library(ISLR2)   # load library
head(Default) |> kable()  # print first six observations
```

We will consider `default` as the response variable.


## Split the data

```{r}
set.seed(427)

default_split <- initial_split(Default, prop = 0.6, strata = default)
default_split

default_train <- training(default_split)
default_test <- testing(default_split)
```

## Summarizing our response variable {.smaller}

```{r}
#| message: false

library(janitor)
Default |> tabyl(default) |> kable()  # class frequencies
default_train |> tabyl(default) |> kable()
default_test |> tabyl(default) |> kable()
```


## Data Types in R {.smaller}

```{r}
Default |> glimpse()
```

-   `fct` = `factor` which is the data type you want to use for categorical data
-   `as_factor` will typically transform things (including numbers) into factors for you
-   `chr` can also be used but `factor`s are better because they store all possible levels for your categorical data
-   `factor`s are helpful for plotting because you can reorder the levels to help you plot things

## K-Nearest Neighbors Classifier

-   Given a value for $K$ and a test data point $x_0$:
    $$P(Y=j | X=x_0)=\dfrac{1}{K} \sum_{x_i \in \mathcal{N}_0} I(y_i = j)$$
    where $\mathcal{N}_0$ is the set of the $K$ "closest" neighbors.
-   For classification: neighbors "vote" for class (unlike in regression
    where predictions are obtained by averaging)
    $$P(Y=j | X=x_0)=\text{Proportion of neighbors in class }j$$


## K-Nearest Neighbors Classifier: Build Model

```{r}
knn_default_fit <- nearest_neighbor(neighbors = 10) |>
  set_engine("kknn") |>
  set_mode("classification") |>
  fit(default ~ balance, data = default_train)   # fit 10-nn model
```

-   Why don't I need to worry about centering and scaling?


## K-Nearest Neighbors Classifier: Predictions

-   `predict` with a categorical response: [documentation](https://parsnip.tidymodels.org/reference/predict.model_fit.html)
-   Two different ways of making predictions

## Predicting a class

```{r}
knn_class_preds <- predict(knn_default_fit, new_data = default_test, type = "class")   # obtain default class label predictions

knn_class_preds |> head() |> kable()
```

##  Predicting a probability

-   Can anyone pick-out what's wrong here? Hint: $k = 10$

```{r}
knn_prob_preds <- predict(knn_default_fit, new_data = default_test, type = "prob")   # obtain predictions as probabilities
knn_prob_preds |> filter(.pred_No*.pred_Yes >0) |> head() |> kable()
```

## I've been lying to you

-   `kknn` actually takes a *weighted average* of the nearest neighbors
    +   I.e. closer observations get more weight
-   To use unweighted KNN need `weight_func = "rectangular"`

## Unweighted KNN

```{r}
knn_default_unw_fit <- nearest_neighbor(neighbors = 10, weight_fun = "rectangular") |>
  set_engine("kknn") |>
  set_mode("classification") |>
  fit(default ~ balance, data = default_train)   # fit 10-nn model

knn_uw_prob_preds <- predict(knn_default_unw_fit, new_data = default_test, type = "prob")   # obtain predictions as probabilities
knn_uw_prob_preds |> filter(.pred_No*.pred_Yes >0) |> head() |> kable()
```

# Logistic Regression

## Why Not Linear Regression? {.smaller}

```{r}
Default_lr <- default_train |> 
  mutate(default_0_1 = if_else(default == "Yes", 1, 0))

lrfit <- linear_reg() |> 
  set_engine("lm") |> 
  fit(default_0_1 ~ balance, data = Default_lr)   # fit SLR

lrfit |> predict(new_data = default_train) |> head() |> kable()
```

## Why Not Linear Regression? {.smaller}

```{r}
#| echo: FALSE
#| fig.align: "center"

ggplot(data = Default_lr, aes(x = balance, y = default_0_1)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  scale_y_continuous(breaks = c(0, 0.5, 1))
```

- Linear regression: does not model probabilities well
  + might produce probabilities less than zero or bigger than one
  + treats increase from 0.41 to 0.5 as same as 0.01 to 0.1 (bad)

## Why Not Linear Regression?

Suppose we have a response,
$$Y=\begin{cases}
1 & \text{if stroke} \\
2 & \text{if drug overdose} \\
3  & \text{if epileptic seizure}
\end{cases}$$

- Linear regression suggests an ordering, and in fact implies that the differences between classes have meaning
  + e.g. drug overdose $-$ stroke $= 1$? `r emo::ji("thinking")`

## Logistic Regression {.smaller}

Consider a one-dimensional binary classification problem:

- Transform the linear model $\beta_0 + \beta_1 \ X$ so that the output is a probability
- Use **logistic** function: $$g(t)=\dfrac{e^t}{1+e^t} \ \ \ \text{for} \ t \in \mathcal{R}$$
- Then: $$p(X)=P(Y=1|X)=g\left(\beta_0 + \beta_1 \ X\right)=\dfrac{e^{\beta_0 + \beta_1 \ X}}{1+e^{\beta_0 + \beta_1 \ X}}$$

## Other important quantities
- **Odds**: $\dfrac{p(x)}{1-p(x)}$
- **Log-Odds (logit)**: $\log\left(\dfrac{p(x)}{1-p(x)}\right) = \beta_0 + \beta_1 \ X$
  + Linear function of predictors

## Logistic Regression


```{r}
#| echo: FALSE
logregfit_orig <- glm(default_0_1 ~ balance, data = Default_lr, family = binomial)   # fit logistic regression model
ggplot(data = Default_lr, aes(x = balance, y = default_0_1)) +
  geom_point() +
  geom_line(aes(y = logregfit_orig$fitted.values), color = "blue") +
  scale_y_continuous(breaks = c(0, 0.5, 1))
```



## Fitting the model

Fitting a logistic regression model with `default` as the response and `balance` as the predictor:

```{r}
logregfit <- logistic_reg() |> 
  set_engine("glm") |> 
  fit(default ~ balance, data = default_train)   # fit logistic regression model

tidy(logregfit) |> kable()  # obtain results
```

## Interpreting Coefficients

- As $X$ increases by 1, the log-odds increase by $\hat{\beta}_1$
  + I.e. probability of default increases but NOT linearly
  + Change in the probability of default due to a one-unit change in balance depends on the current balance value

## Interpreting Coefficients

```{r}
#| echo: FALSE
x0 <- 1000
x1 <- 1500
x2 <- 2000
y0 <- predict(logregfit_orig, newdata = data.frame(balance = x0), type = "response")
y1 <- predict(logregfit_orig, newdata = data.frame(balance = x1), type = "response")
y2 <- predict(logregfit_orig, newdata = data.frame(balance = x2), type = "response")
ggplot(data = Default_lr, aes(x = balance, y = default_0_1)) +
  geom_point() +
  geom_vline(xintercept = x0, color = "red", alpha = 1, linetype = "dotted") +
  geom_vline(xintercept = x1, color = "red", alpha = 1, linetype = "dotted") +
  geom_vline(xintercept = x2, color = "red", alpha = 1, linetype = "dotted") +
  geom_hline(yintercept = y0, color = "blue", alpha = 0.5) +
  geom_hline(yintercept = y1, color = "blue", alpha = 0.5) +
  geom_hline(yintercept = y2, color = "blue", alpha = 0.5) +
 geom_segment(aes(x = 1250, y = y0, xend = 1250, yend = y1),
                  arrow = arrow(length = unit(0.1, "cm"), type = "closed", ends="both"), color = "blue") +
   geom_segment(aes(x = 1750, y = y1, xend = 1750, yend = y2),
                  arrow = arrow(length = unit(0.1, "cm"), type = "closed", ends="both"), color = "blue") +
  geom_line(aes(y = logregfit_orig$fitted.values)) +
  scale_y_continuous(breaks = c(0, 0.5, 1)) +
  ylab("Probility of Default")
```

## Making predictions: Theory {.smaller}

For `balance`=\$700,

:::{.incremental}
- $$\hat{p}(X)=\dfrac{e^{\hat{\beta}_0+\hat{\beta}_1 X}}{1+e^{\hat{\beta}_0+\hat{\beta}_1 X}}=\dfrac{e^{-10.69 + (0.005533 \times 700)}}{1+e^{-10.69 + (0.005533 \times 700)}}=0.0011$$
-  $$\textbf{Odds}(X) = \dfrac{\hat{p}(X)}{1-\hat{p}(X)} = \dfrac{0.0011}{1-0.0011}\approx 0.0011$$
- $$\textbf{Log-Odds}(X)=\log\left(\dfrac{\hat{p}(X)}{1-\hat{p}(X)}\right) = \log(0.0011) = -6.80$$
:::

## Making predictions in R

```{r}
predict(logregfit, new_data = tibble(balance = 700), type = "class") |> kable()   # obtain class predictions
predict(logregfit, new_data = tibble(balance = 700), type = "raw") |> kable()   # obtain log-odds predictions
predict(logregfit, new_data = tibble(balance = 700), type = "prob") |> kable()  # obtain probability predictions
```