---
title: 'MATH 427: Evaluating Classification Models'
author: Eric Friedlander
footer: "[🔗 MAT 427 - Spring 2025 -  Schedule](https://mat427sp25.netlify.app/schedule)"
logo: "../images/logo.png"
format: 
  revealjs:
    theme: slides.scss
    multiplex: false
    transition: fade
    slide-number: false
    incremental: false 
    chalkboard: true
execute:
  freeze: auto
  echo: true
knitr:
  opts_chunk: 
    R.options:      
    width: 200
---

## Tips on Gradient Descent HW

-   Make step size small!
-   May take a while to converge
-   Try adaptive step size (i.e. backtracking)
-   Clarification on stopping criteria
    +   Set tolerance
    +   Stop when distance from gradient to $(0, 0)$ is below tolerance


## Computational Set-Up

```{r}
library(tidyverse)
library(tidymodels)
library(gridExtra)
library(janitor) # for next contingency tables
library(kableExtra)
library(ISLR2)

tidymodels_prefer()
```



## Default Dataset {.smaller}

::: columns
::: column
A simulated data set containing information on ten thousand customers. The aim here is to predict which customers will default on their credit card debt.

```{r}
#| message: FALSE
head(Default) |> kable()  # print first six observations
```
:::

::: column
**Response Variable**: `default`

```{r}
Default |> 
  tabyl(default) |>  # class frequencies
  kable()           # Make it look nice
```
:::
:::

## Split the data

```{r}
set.seed(427)

default_split <- initial_split(Default, prop = 0.6, strata = default)
default_split

default_train <- training(default_split)
default_test <- testing(default_split)
```

## [K-Nearest Neighbors Classifier: Build Model]{.r-fit-text}

-   **Response** ($Y$): `default`
-   **Predictor** ($X$): `balance`

```{r}
knnfit <- nearest_neighbor(neighbors = 10) |> 
  set_engine("kknn") |> 
  set_mode("classification") |>  
  fit(default ~ balance, data = Default)   # fit 10-nn model
```

## [K-Nearest Neighbors Classifier: Predictions]{.r-fit-text} {.smaller}

::: panel-tabset
## Class labels

```{r}
predict(knnfit, new_data = Default, type = "class") |> head() |> kable()   # obtain predictions as classes
```

## Probabilities

-   Predicts class w/ maximum probability

```{r}
predict(knnfit, new_data = Default, type = "prob") |> head() |> kable() # obtain predictions as probabilities
```
:::

## Fitting a logistic regression

Fitting a logistic regression model with `default` as the response and `balance` as the predictor:

```{r}
logregfit <- logistic_reg() |> 
  set_engine("glm") |> 
  fit(default ~ balance, data = default_train)   # fit logistic regression model

tidy(logregfit) |> kable()  # obtain results
```

## Making predictions in R

::: panel-tabset

## Class Labels

```{r}
predict(logregfit, new_data = tibble(balance = 700), type = "class") |> kable()   # obtain class predictions
```

## Log-Odds

```{r}
predict(logregfit, new_data = tibble(balance = 700), type = "raw") |> kable()   # obtain log-odds predictions
```

## Probabilities

```{r}
predict(logregfit, new_data = tibble(balance = 700), type = "prob") |> kable()  # obtain probability predictions
```

:::


# Assessing Performance of Classifiers

## Binary Classifiers

-   Start with binary classification scenarios
-   With binary classification, designate one category as "Success/Positive" and the other as "Failure/Negative"
    +   If relevant to your problem: "Positive" should be the thing you're trying to predict/care more about
    +   Note: "Positive" $\neq$ "Good"
    +   For `default`: "Yes" is Positive
-   Some metrics weight "Positives" more and viceversa

## Confusion Matrix

|                                  | Actual Positive/Event | Actual Negative/Non-event |
|:--------------------------------:|:---------------------:|:-------------------------:|
|   **Predicted Positive/Event**   |     True Positive (TP)     |      False Positive  (FP)     |
| **Predicted Negative/Non-event** |    False Negative (FN)    |       True Negative  (TN)     |

## Adding predictions to tibble

```{r}
default_test_wpreds <- default_test |> 
  mutate(
    knn_preds = predict(knnfit, new_data = default_test, type = "class")$.pred_class,
    logistic_preds = predict(logregfit, new_data = default_test, type = "class")$.pred_class
  )

default_test_wpreds |> head() |> kable()
```

## KNN: Confusion Matrix

```{r}
default_test_wpreds |>
  conf_mat(truth = default, estimate = knn_preds)
```

## KNN: Confusion Matrix (Sexy)

```{r}
default_test_wpreds |>
  conf_mat(truth = default, estimate = knn_preds) |> 
  autoplot("heatmap")
```

## Logistic Regression: Confusion Matrix

```{r}
default_test_wpreds |>
  conf_mat(truth = default, estimate = logistic_preds) |> 
  autoplot(type = "heatmap")
```

## Classification Metrics {.smaller}

-   Accuracy: proportion of your classes that are correct $$(TP + TN)/Total$$
-   Recall/Sensitivity: proportion of true positives correct (true positive rate) $$TP/(TP+FN)$$
-   Precision/Positive Predictive Value (PPV): proportion of predicted positive that are correct $$TP/(TP+FP)$$
-   Specificity: proportion of true negatives correct (true negative rate) $$TN/(TN+FP)$$
-   Negative Predictive Value (NPV): proportion of predicted negatives that are correct $$TN/(TN+FN)$$

## KNN: Performance {.smaller}

:::: columns
::: column
```{r}
default_test_wpreds |>
  conf_mat(truth = default, estimate = knn_preds) |> 
  autoplot("heatmap")
```
:::

::: column
-   Accuracy: $(3854+49)/4000 = .976 = 97.6\%$
-   Recall/Sensitivity: $49/(49+80) = 0.380 = 38.0\%$
-   Precision/Positive Predictive Value (PPV): $49/(49+17) = .742 = 74.2\%$
-   Specificity: $3854/(3854+17) = 0.996 = 99.6\%$
-   Negative Predictive Value (NPV): $3854/(3854+80) = 98.0$
:::
::::

## Logistic Regression: Performance {.smaller}

:::: columns
::: column
```{r}
default_test_wpreds |>
  conf_mat(truth = default, estimate = logistic_preds) |> 
  autoplot("heatmap")
```
:::

::: column
Compute the following and write your answers on the board:

-   Accuracy
-   Recall/Sensitivity
-   Precision/Positive Predictive Value (PPV)
-   Specificity
-   Negative Predictive Value (NPV)
:::
::::

## Performance Metrics with `yardstick`

-   `yardstick` is a package that ships with `tidymodels` meant for model evaluation
-   Typical syntax: `metricname(data, truth, estimate, ...)`
    +   Bind original data with predicted observations
    +   Put true response in for `truth` and predicted values in for `estimate`
    
## Logistic Regression: Accuracy {.smaller}


```{r}
default_test_wpreds |> 
  accuracy(truth = default, estimate = logistic_preds) |> 
  kable()
```

## Two More Metrics {.smaller}

-   Matthews correlation coefficient (MCC): similar to $R^2$ but for classification
$$\frac{TP\times TN - FP \times FN}{\sqrt{(TP + FP)(TP+FN)(TN+FP)(TN+FN)}}$$
    +   Good for imbalanced data
    +   Considers both positives and negatives
-   F-Measure: harmonic mean of recall and precision
$$\frac{2}{recall^{-1} + precision^{-1}} = \frac{2TP}{2TP+FP+FN}$$
    +   Focuses more on positives
    +   bad of imbalanced data

## Metric Sets

```{r}
binary_metrics <- metric_set(accuracy, recall, precision, specificity,
                             npv, mcc, f_meas)
```

-   Can apply this to compute a bunch of metrics


## KNN: Performance {.smaller}

```{r}
default_test_wpreds |> 
  binary_metrics(truth = default, estimate = knn_preds, event_level = "second") |> 
  kable()
```

## Logistic Regression: Performance {.smaller}

```{r}
default_test_wpreds |> 
  binary_metrics(truth = default, estimate = logistic_preds, event_level = "second") |> 
  kable()
```

## Discussion

-   For each of the following metrics, brainstorm a situation in which that metric is probably the most important:
    +   Recall
    +   Precision
    +   Accuracy