---
title: 'MATH 427: Workflow Sets and Feature Selection'
author: Eric Friedlander
footer: "[🔗 MAT 427 - Spring 2025 -  Schedule](https://mat427sp25.netlify.app/schedule)"
logo: "../images/logo.png"
format: 
  revealjs:
    theme: slides.scss
    multiplex: false
    transition: fade
    slide-number: false
    incremental: false 
    chalkboard: true
execute:
  freeze: auto
  echo: true
  cache: true
knitr:
  opts_chunk: 
    R.options:      
    width: 200
---

## Computational Set-Up

```{r}
library(tidyverse)
library(tidymodels)
library(knitr)
library(janitor) # for contingency tables
library(ISLR2)
library(readODS)

tidymodels_prefer()

set.seed(427)
```


# Workflow Sets in R

## Data: Different Ames Housing Prices {.smaller}

Goal: Predict `Sale_Price`.

```{r}
ames <- read_rds("../data/AmesHousing.rds")
ames |> glimpse()
```


```{r}
#| echo: FALSE
ames <- ames |> 
  mutate(Garage_Type = if_else(Garage_Type == "No_Garage", NA, Garage_Type),
        rand_num2 = runif(881, 0, 1),
        Year_Built = if_else(rand_num2 < 0.05, NA, Year_Built)) |> 
  select(-rand_num2)
```

## Clean Data Set

```{r}
ames <- ames |> 
  mutate(Overall_Qual = factor(Overall_Qual, levels = c("Very_Poor", "Poor", 
                                                        "Fair", "Below_Average",
                                                        "Average", "Above_Average", 
                                                        "Good", "Very_Good",
                                                        "Excellent", "Very_Excellent")),
         Garage_Type = if_else(is.na(Garage_Type), "No_Garage", Garage_Type),
         Garage_Type = as_factor(Garage_Type)
         )
```

## Initial Data Split

```{r}
data_split <- initial_split(ames, strata = "Sale_Price")
ames_train <- training(data_split)
ames_test  <- testing(data_split)
```

## Define Folds

```{r}
ames_folds <- vfold_cv(ames_train, v = 10, repeats = 10)
ames_folds
```

## Define Model(s)

```{r}
lm_model <- linear_reg() |> 
  set_engine("lm")

knn5_model <- nearest_neighbor(neighbors = 5) |>
  set_engine("kknn") |>
  set_mode("regression")

knn10_model <- nearest_neighbor(neighbors = 10) |>
  set_engine("kknn") |>
  set_mode("regression")
```

## Define Preprocessing: Linear regression

```{r}
lm_knnimpute <- recipe(Sale_Price ~ ., data = ames_train) |> 
  step_nzv(all_predictors()) |>  # remove zero or near-zero variable predictors
  step_impute_knn(Year_Built, Gr_Liv_Area) |>  # impute missing values in Overall_Qual and Year_Built
  step_integer(Overall_Qual) |> # convert Overall_Qual into ordinal encoding
  step_other(all_nominal_predictors(), threshold = 0.01, other = "Other") |> # lump all categories with less than 1% representation into a category called Other for each variable
  step_dummy(all_nominal_predictors(), one_hot = FALSE) |>  # in general use one_hot unless doing linear regression
  step_corr(all_numeric_predictors(), threshold = 0.5) |> # remove highly correlated predictors
  step_lincomb(all_numeric_predictors()) # remove variables that have exact linear combinations
```

## Define Preprocessing: Linear regression

```{r}
lm_meanimpute <- recipe(Sale_Price ~ ., data = ames_train) |> 
  step_nzv(all_predictors()) |>  # remove zero or near-zero variable predictors
  step_impute_mean(Year_Built, Gr_Liv_Area) |>  # impute missing values in Overall_Qual and Year_Built
  step_integer(Overall_Qual) |> # convert Overall_Qual into ordinal encoding
  step_other(all_nominal_predictors(), threshold = 0.01, other = "Other") |> # lump all categories with less than 1% representation into a category called Other for each variable
  step_dummy(all_nominal_predictors(), one_hot = FALSE) |>  # in general use one_hot unless doing linear regression
  step_corr(all_numeric_predictors(), threshold = 0.5) |> # remove highly correlated predictors
  step_lincomb(all_numeric_predictors()) # remove variables that have exact linear combinations
```

## Define Preprocessing: Linear regression

```{r}
lm_medianimpute <- recipe(Sale_Price ~ ., data = ames_train) |> 
  step_nzv(all_predictors()) |>  # remove zero or near-zero variable predictors
  step_impute_median(Year_Built, Gr_Liv_Area) |>  # impute missing values in Overall_Qual and Year_Built
  step_integer(Overall_Qual) |> # convert Overall_Qual into ordinal encoding
  step_other(all_nominal_predictors(), threshold = 0.01, other = "Other") |> # lump all categories with less than 1% representation into a category called Other for each variable
  step_dummy(all_nominal_predictors(), one_hot = FALSE) |>  # in general use one_hot unless doing linear regression
  step_corr(all_numeric_predictors(), threshold = 0.5) |> # remove highly correlated predictors
  step_lincomb(all_numeric_predictors()) # remove variables that have exact linear combinations
```

## Define Preprocessing: KNN

```{r}
knn_preproc1 <- recipe(Sale_Price ~ ., data = ames_train) |> 
  step_nzv(all_predictors()) |>  # remove zero or near-zero variable predictors
  step_impute_knn(Year_Built, Gr_Liv_Area) |>  # impute missing values in Overall_Qual and Year_Built
  step_integer(Overall_Qual) |> # convert Overall_Qual into ordinal encoding
  step_other(all_nominal_predictors(), threshold = 0.01, other = "Other") |> # lump all categories with less than 1% representation into a category called Other for each variable
  step_dummy(all_nominal_predictors(), one_hot = TRUE) |>  # in general use one_hot unless doing linear regression
  step_nzv(all_predictors()) |> 
  step_normalize(all_numeric_predictors())
```

## Define Preprocessing: KNN

```{r}
knn_preproc2 <- recipe(Sale_Price ~ ., data = ames_train) |> 
  step_nzv(all_predictors()) |>  # remove zero or near-zero variable predictors
  step_impute_mean(Year_Built, Gr_Liv_Area) |>  # impute missing values in Overall_Qual and Year_Built
  step_integer(Overall_Qual) |> # convert Overall_Qual into ordinal encoding
  step_other(all_nominal_predictors(), threshold = 0.01, other = "Other") |> # lump all categories with less than 1% representation into a category called Other for each variable
  step_dummy(all_nominal_predictors(), one_hot = TRUE) |>  # in general use one_hot unless doing linear regression
  step_nzv(all_predictors()) |> 
  step_normalize(all_numeric_predictors())
```

## Define Preprocessing: KNN

```{r}
knn_preproc3 <- recipe(Sale_Price ~ ., data = ames_train) |> 
  step_nzv(all_predictors()) |>  # remove zero or near-zero variable predictors
  step_impute_median(Year_Built, Gr_Liv_Area) |>  # impute missing values in Overall_Qual and Year_Built
  step_integer(Overall_Qual) |> # convert Overall_Qual into ordinal encoding
  step_other(all_nominal_predictors(), threshold = 0.01, other = "Other") |> # lump all categories with less than 1% representation into a category called Other for each variable
  step_dummy(all_nominal_predictors(), one_hot = TRUE) |>  # in general use one_hot unless doing linear regression
  step_nzv(all_predictors()) |> 
  step_normalize(all_numeric_predictors())
```

## Workflow Sets

-   Input `list`s of models and `recipe`s
-   If `cross = TRUE` will try out all combinations

## Create lists

```{r}
knn_preprocessors <- list(
  knn_knn_impute = knn_preproc1,
  knn_mean_impute = knn_preproc2,
  knn_median_imput = knn_preproc3
)

knn_models <- list(
  knn5 = knn5_model,
  knn10 = knn10_model
)
```

## Create lists

```{r}
lm_preprocessors <- list(
  lm_knn_impute = lm_knnimpute,
  lm_mean_impute = lm_meanimpute,
  lm_median_imput = lm_medianimpute
)

lm_models <- list(
  lm_model = lm_model
)
```

## Define Workflow Sets {.smaller}

```{r}
knn_models <- workflow_set(knn_preprocessors, knn_models, cross = TRUE)
lm_models <-  workflow_set(lm_preprocessors, lm_models, cross = TRUE)
all_models <- lm_models |> 
  bind_rows(knn_models)
  
all_models
```

## Define Metrics

```{r}
ames_metrics <- metric_set(rmse, rsq)
```

## Fit Resamples {.smaller}

```{r}
all_fits <- all_models |> 
  workflow_map("fit_resamples",
               resamples = ames_folds,
               metrics = ames_metrics)
```

## View Metrics

```{r}
collect_metrics(all_fits) |> 
  filter(.metric == "rmse") |> 
  kable()
```

## View Metrics

```{r}
collect_metrics(all_fits) |> 
  filter(.metric == "rsq") |> 
  kable()
```

## Plotting Results

```{r}
library(ggrepel)
autoplot(all_fits, metric = "rmse") +
  geom_text_repel(aes(label = wflow_id), nudge_x = 1/8, nudge_y = 1/100) +
  theme(legend.position = "none")
```

## Plotting Results

```{r}
autoplot(all_fits, metric = "rsq") +
  geom_text_repel(aes(label = wflow_id), nudge_x = 1/8, nudge_y = 1/100) +
  theme(legend.position = "none")
```




# Feature Selection for Linear Regression

## What is feature selection? {.smaller}

-   How do we choose what variables to include in our model?
-   Up to now... include all of them... probably not the best
-   Advantage of linear regression: interpretability
-   Including every feature decreases interpretability
-   Reasons for feature selection:
    +   Improve model performance
    +   Improve model interpretability
-   Parsimony: simpler models are called more *parsimonious*
-   Occam's Razor: more parsimonious models are better than less parsimonious models, holding all else constant

## Types of feature selection

-   Subset selection: Forward/Backward/Best-Subset Selection
-   Shrinkage-based methods: LASSO and Ridge Regression
-   Dimension reduction: consider linear combinations of predictors

# Subset Selection

## Exercise

-   With your group, write out the steps for the following algorithms on the board
  +   Group 1: Forward selection
  +   Group 2: Backward elimination
  +   Group 3: Step-wise selection
  +   Group 4: Best-subset selection

## Subset Selection in R {.smaller}

-   `tidymodels` does not have an implementation for any subset selection techniques
-   regularization (shrinkage-based) methods almost always perform better
-   [`colino` package](https://stevenpawley.github.io/colino/) provides `tidymodels` implementation
-   Other options
    +   `caret` package
    +   `olsrr` and `blorr` packages if you don't care about cross-validation
    +   implement yourself
    
## Feature Selection in R

-   When creating your recipe, don't need to always include all variables in your `recipe`:

```{r}
#| eval: FALSE
int_recipe <- recipe(pred ~ var1 + var2 + var1*var2, data = training_data) |> 
  step_x(...)
```

## Re-using Recipe but changing formula

```{r}
#| eval: FALSE
noint_recipe2 <- new_recipe |> 
  remove_formula() |> 
  add_formula(pred ~ var1 + var2)
```