---
title: 'MATH 427: Feature Selection and Regularization'
author: Eric Friedlander
footer: "[🔗 MAT 427 - Spring 2025 -  Schedule](https://mat427sp25.netlify.app/schedule)"
logo: "../images/logo.png"
format: 
  revealjs:
    theme: slides.scss
    multiplex: false
    transition: fade
    slide-number: false
    incremental: false 
    chalkboard: true
execute:
  freeze: auto
  echo: true
  cache: true
knitr:
  opts_chunk: 
    R.options:      
    width: 200
---

## Computational Set-Up

```{r}
library(tidyverse)
library(tidymodels)
library(knitr)
library(fivethirtyeight) # for candy rankings data

tidymodels_prefer()

set.seed(427)
```

## Data: Candy {.smaller}

The data for this lecture comes from the article FiveThirtyEight [*The Ultimate Halloween Candy Power Ranking*](https://fivethirtyeight.com/features/the-ultimate-halloween-candy-power-ranking) by Walt Hickey. To collect data, Hickey and collaborators at FiveThirtyEight set up an experiment people could vote on a series of randomly generated candy matchups (e.g. Reese's vs. Skittles). Click [here](http://walthickey.com/2017/10/18/whats-the-best-halloween-candy/) to check out some of the match ups.

The data set contains 12 characteristics and win percentage from 85 candies in the experiment.

## Data: Candy

```{r}
glimpse(candy_rankings)
```

## Data Cleaning

```{r}
candy_rankings_clean <- candy_rankings |> 
  select(-competitorname) |> 
  mutate(sugarpercent = sugarpercent*100, # convert proportions into percentages
         pricepercent = pricepercent*100, # convert proportions into percentages
         across(where(is.logical), ~ factor(.x, levels = c("FALSE", "TRUE")))) # convert logicals into factors
```

## Data Cleaning

```{r}
glimpse(candy_rankings_clean)
```

## Data Splitting

```{r}
candy_split <- initial_split(candy_rankings_clean, prop = 0.75, strata = winpercent)
candy_train <- training(candy_split)
candy_test <- testing(candy_split)
```


# Feature Selection for Linear Regression

## What is feature selection? {.smaller}

-   How do we choose what variables to include in our model?
-   Up to now... include all of them... probably not the best
-   Advantage of linear regression: interpretability
-   Including every feature decreases interpretability
-   Reasons for feature selection:
    +   Improve model performance
    +   Improve model interpretability
-   Parsimony: simpler models are called more *parsimonious*
-   Occam's Razor: more parsimonious models are better than less parsimonious models, holding all else constant

## Types of feature selection

-   Subset selection: Forward/Backward/Best-Subset Selection
-   Shrinkage-based methods: LASSO and Ridge Regression
-   Dimension reduction: consider linear combinations of predictors

# Subset Selection

## Exercise

-   With your group, write out the steps for the following algorithms on the board
    +   Group 1: Forward selection
    +   Group 2: Backward elimination
    +   Group 3: Step-wise selection
    +   Group 4: Best-subset selection

## Subset Selection in R {.smaller}

-   `tidymodels` does not have an implementation for any subset selection techniques
-   regularization (shrinkage-based) methods almost always perform better
-   [`colino` package](https://stevenpawley.github.io/colino/) provides `tidymodels` implementation
-   Other options
    +   `caret` package
    +   `olsrr` and `blorr` packages if you don't care about cross-validation
    +   implement yourself
    
## Feature Selection in R

-   When creating your recipe, don't need to always include all variables in your `recipe`:

```{r}
#| eval: FALSE
int_recipe <- recipe(resp ~ var1 + var2 + var1*var2, data = training_data) |> 
  step_x(...)
```

## Re-using Recipe but changing formula

```{r}
#| eval: FALSE
noint_recipe2 <- new_recipe |> 
  remove_formula() |> 
  add_formula(resp ~ var1 + var2)
```

# Shrinkage/Penalized/Regularization Methods

## Question

-   What criteria do we use to fit a linear regression model? Write down an expression with $\hat{\beta_i}$'s, $x_{ij}$'s, and $y_j$'s in it.

## OLS {.smaller}

-   Ordinary Least Squares Regression:

$$
\begin{aligned}
\hat{\beta} =\underset{\hat{\beta}_0,\ldots, \hat{\beta}_p}{\operatorname{argmin}}SSE(\hat{\beta}) &= \underset{\hat{\beta}_0,\ldots, \hat{\beta}_p}{\operatorname{argmin}}\sum_{j=1}^n(y_j-\hat{y}_j)^2\\
&=\underset{\hat{\beta}_0,\ldots, \hat{\beta}_p}{\operatorname{argmin}}\sum_{j=1}^n(y_j-\hat{\beta}_0-\hat{\beta}_1x_{1j} - \cdots - \hat{\beta}_px_{pj})^2
\end{aligned}
$$

-   $\hat{\beta} = (\hat{\beta}_0,\ldots, \hat{\beta}_p)$ is the vector of all my coefficients
-   $\operatorname{argmin}$ is a function (operator) that returns the *arguments* that minimize the quantity it's being applied to

## Ridge Regression {.smaller}

$$
\begin{aligned}
\hat{\beta} &=\underset{\hat{\beta}_0,\ldots, \hat{\beta}_p}{\operatorname{argmin}} \left(SSE(\hat{\beta}) + \lambda\|\hat{\beta}\|_2^2\right) \\
&= \underset{\hat{\beta}_0,\ldots, \hat{\beta}_p}{\operatorname{argmin}}\left(\sum_{j=1}^n(y_j-\hat{y}_j)^2 + \lambda\sum_{i=1}^p \hat{\beta}_i^2\right)\\
&=\underset{\hat{\beta}_0,\ldots, \hat{\beta}_p}{\operatorname{argmin}}\left(\sum_{j=1}^n(y_j-\hat{\beta}_0-\hat{\beta}_1x_{1j} - \cdots - \hat{\beta}_px_{pj})^2 + \lambda\sum_{i=1}^p \hat{\beta}_i^2\right)
\end{aligned}
$$

-   $\lambda$ is called a "tuning parameter" and is not estimated from the data (more on this later)
-   Idea: penalize large coefficients
-   If we penalize large coefficients, what's going to happen to to our estimated coefficients? [[THEY SHRINK!]{.span style="color:red;"}]{.fragment .fade-in}
-   $\|\cdot\|_2$ is called the *$L_2$-norm*

## But... but... but why?!? {.smaller}

::: incremental
-   Recall the Bias-Variance Trade-Off
-   Our reducible error can partitioned into:
    +   Bias: how much $\hat{f}$ misses $f$ by *on average*
    +   Variance: how much $\hat{f}$ moves around from sample to sample
-   Ridge: increase bias a little bit in exchange for large decrease in variance
-   As we increase $\lambda$ do we increase or decrease the penalty for large coefficients?
-   As we increase $\lambda$ do we increase or decrease the *flexibility* of our model?
:::

## LASSO {.smaller}

$$
\begin{aligned}
\hat{\beta} &=\underset{\hat{\beta}_0,\ldots, \hat{\beta}_p}{\operatorname{argmin}} \left(SSE(\hat{\beta}) + \lambda\|\hat{\beta}\|_1\right) \\
&= \underset{\hat{\beta}_0,\ldots, \hat{\beta}_p}{\operatorname{argmin}}\left(\sum_{j=1}^n(y_j-\hat{y}_j)^2 + \lambda\sum_{i=1}^p |\hat{\beta}_i|\right)\\
&=\underset{\hat{\beta}_0,\ldots, \hat{\beta}_p}{\operatorname{argmin}}\left(\sum_{j=1}^n(y_j-\hat{\beta}_0-\hat{\beta}_1x_{1j} - \cdots - \hat{\beta}_px_{pj})^2 + \lambda\sum_{i=1}^p |\hat{\beta}_i|\right)
\end{aligned}
$$

-   LASSO: **L**east **A**bsolute **S**hrinkage and **S**election **O**perator
-   $\lambda$ is called a "tuning parameter" and is not estimated from the data (more on this later)
-   Idea: penalize large coefficients
-   If we penalize large coefficients, what's going to happen to to our estimated coefficients [THEY SHRINK!]{.span style="color:red;"}
-   $\|\cdot\|_1$ is called the *$L_1$-norm*

## Question

-   What should happen to our coefficients as we *increase* $\lambda$?

## Ridge vs. LASSO

::: incremental
-   Ridge has a *closed-form solution*... how might we calculate it?
-   Ridge has some nice linear algebra properties that makes it EXTREMELY FAST to fit
-   LASSO has no *closed-form solution*... why?
-   LASSO coefficients estimated *numerically*... how?
    +   Gradient descent works (kind of) but something called **coordinate descent** is typically better
-   MOST IMPORTANT PROPERTY OF LASSO: it *induces sparsity* while Ridge does not
:::

## Sparsity in Applied Math {.smaller}

-   **sparse** typically means "most things are zero"
-   Example: sparse matrices are matrices where most entries are zero
    +   for large matrices this can provide HUGE performance gains
-   LASSO *induces sparsity* by setting most of the parameter estimates to zero
    +   this means it fits the model and does feature selection SIMULTANEOUSLY
-   Let's do some board work to see why this is...

