---
title: 'MATH 427: Regularization & Model Tuning'
author: Eric Friedlander
footer: "[🔗 MAT 427 - Spring 2025 -  Schedule](https://mat427sp25.netlify.app/schedule)"
logo: "../images/logo.png"
format: 
  revealjs:
    theme: slides.scss
    multiplex: false
    transition: fade
    slide-number: false
    incremental: false 
    chalkboard: true
execute:
  freeze: auto
  echo: true
  cache: true
knitr:
  opts_chunk: 
    R.options:      
    width: 200
---


## Computational Set-Up

```{r}
library(tidyverse)
library(tidymodels)
library(knitr)
library(fivethirtyeight) # for candy rankings data

tidymodels_prefer()

set.seed(427)
```

## Data: Candy {.smaller}

The data for this lecture comes from the article FiveThirtyEight [*The Ultimate Halloween Candy Power Ranking*](https://fivethirtyeight.com/features/the-ultimate-halloween-candy-power-ranking) by Walt Hickey. To collect data, Hickey and collaborators at FiveThirtyEight set up an experiment people could vote on a series of randomly generated candy match-ups (e.g. Reese's vs. Skittles). Click [here](http://walthickey.com/2017/10/18/whats-the-best-halloween-candy/) to check out some of the match ups.

The data set contains 12 characteristics and win percentage from 85 candies in the experiment.

## Data: Candy

```{r}
glimpse(candy_rankings)
```

## Data Cleaning

```{r}
candy_rankings_clean <- candy_rankings |> 
  select(-competitorname) |> 
  mutate(sugarpercent = sugarpercent*100, # convert proportions into percentages
         pricepercent = pricepercent*100, # convert proportions into percentages
         across(where(is.logical), ~ factor(.x, levels = c("FALSE", "TRUE")))) # convert logicals into factors
```

## Data Cleaning

```{r}
glimpse(candy_rankings_clean)
```

## Data Splitting

```{r}
candy_split <- initial_split(candy_rankings_clean, strata = winpercent)
candy_train <- training(candy_split)
candy_test <- testing(candy_split)
```

# Model Tuning

## Tuning Parameters

-   **Tuning Parameters** or **Hyperparameters** are parameters that cannot (or should not) be estimated when a model is being trained
-   These parameters control something about the learning process and changing them will result in a different model when fit to the full training data (i.e. after cross-validation)
-   Frequently: tuning parameters control model complexity
-   Example: $\lambda$ in LASSO and Ridge regression
-   Today: How to choose our tuning parameters?

## Question

-   Which of the following are tuning parameters:
    +   $\beta_0$: the intercept of linear regression
    +   $k$ in KNN
    +   step size in gradient descent
    +   The number of folds in cross-validation
    +   Type of distance to use in KNN (i.e. rectangular vs. Gower's vs. weighted etc)

## Basic Idea

-   Use CV to try out a bunch of different tuning parameters and choose the "best" one
-   How do we choose which tuning parameters to try?
-   Two general approaches:
    +   Grid Search
    +   Iterative Search
    
# Grid Search
    
## Grid Search

-   Create a grid of tuning parameters and try out each combination
-   Types of grids:
    +   Regular Grid: tuning parameter values are spaced *deterministically* using a linear or logarithmic scale and all combinations of parameters are used (mostly what you want to use in this class)
    +   Irregular Grids: tuning parameter values are chosen *stochastically*
        -   Use when you have A LOT of parameters
    
## Grid Search in R

-   Take advantage of package `dials` which is part of the `tidyverse`
    +   Set every tuning variable equal to `tune()`
    
## LASSO and Ridge in R

```{r}
ols <- linear_reg() |> 
  set_engine("lm")

ridge <- linear_reg(mixture = 0, penalty = tune()) |> # penalty set's our lambda
  set_engine("glmnet")

lasso <- linear_reg(mixture = 1, penalty = tune()) |> # penalty set's our lambda
  set_engine("glmnet")
```
    
## Create Recipe

-   Note: no tuning variables in this case

```{r}
lm_preproc <- recipe(winpercent ~ . , data = candy_train) |> 
  step_dummy(all_nominal_predictors()) |> 
  step_normalize(all_numeric_predictors())
```

## Create workflows

```{r}
generic_wf <- workflow() |> add_recipe(lm_preproc)
ols_wf <- generic_wf |> add_model(ols)
ridge_wf <- generic_wf |> add_model(ridge)
lasso_wf <- generic_wf |> add_model(lasso)
```

## Create Metric Set

```{r}
candy_metrics <- metric_set(rmse, rsq)
```

## Create Folds

```{r}
# Since sample size is so small, not using stratification
candy_folds <- vfold_cv(candy_train, v = 2, repeats = 10)
```

## Grid Search in R

-   Take advantage of package `dials` which is part of the `tidyverse`
    +   Set every tuning variable equal to `tune()`
    +   Generate grid for hyperparameters

## Generate Grid

```{r}
# note that dials treats penalty on a log10 scale
penalty_grid <- grid_regular(penalty(range = c(-10, 2)), # I had to play around with these 
                             levels = 10)
```

:::: columns
::: column
```{r}
penalty_grid |> head() |>  kable(digits = 11, format.args = list(scientific = TRUE))
```
:::

::: column
```{r}
penalty_grid |> tail() |>  kable()
```
:::
::::


## Regular Grid Search in R

-   Take advantage of package `dials` which is part of the `tidyverse`
    +   Set every tuning variable equal to `tune()`
    +   Generate grid for hyperparameters using `grid_regular`
    +   Tune your model: fit all hyperparameter combination on resamples using `tune_grid`

## Tune Models

```{r}
tuning_ridge_results <- tune_grid(
  ridge_wf,
  resamples= candy_folds,
  grid = penalty_grid
)

tuning_lasso_results <- tune_grid(
  lasso_wf,
  resamples = candy_folds,
  grid = penalty_grid
)
```

## Regular Grid Search in R

-   Take advantage of package `dials` which is part of the `tidyverse`
    +   Set every tuning variable equal to `tune()`
    +   Generate grid for hyperparameters using `grid_regular`
    +   Tune your model: fit all hyperparameter combination on resamples using `tune_grid`
    +   Visualize and Choose final model

## Visualizing Results: Ridge

```{r}
autoplot(tuning_ridge_results)
```

## Selecting Best Model: Ridge

:::: columns
::: column

Best RMSE

```{r}
best_rmse_ridge <- tuning_ridge_results |> 
  select_best(metric = "rmse")
best_rmse_ridge |> kable(digits = 2)
```
:::

::: column
Best $R^2$
```{r}
best_rsq_ridge <- tuning_ridge_results |> 
  select_best(metric = "rsq",)
best_rsq_ridge |> kable(digits = 2)
```
:::
::::

-   Which should be use?

## Finalize Model {.smaller}

-   RMSE estimate is less flexible and seems to be sacrificing less $R^2$ than the opposite

```{r}
ridge_rmse_final <- finalize_workflow(ridge_wf, best_rmse_ridge)
ridge_rmse_fit <- fit(ridge_rmse_final, data = candy_train)
tidy(ridge_rmse_fit) |> 
  kable()
```

## Visualizing Results: LASSO

```{r}
autoplot(tuning_lasso_results)
```

## Selecting Best Model: LASSO

:::: columns
::: column

Best RMSE

```{r}
best_rmse_lasso <- tuning_lasso_results |> 
  select_best(metric = "rmse")
best_rmse_lasso |> kable(digits = 2)
```
:::

::: column
Best $R^2$
```{r}
best_rsq_lasso <- tuning_lasso_results |> 
  select_best(metric = "rsq")
best_rsq_lasso |> kable(digits = 2)
```
:::
::::

-   Which should be use?

## Finalize Model {.smaller}

-   RMSE estimate is less flexible and seems to be sacrificing less $R^2$ than the opposite

```{r}
lasso_rmse_final <- finalize_workflow(lasso_wf, best_rmse_lasso)
lasso_rmse_fit <- fit(lasso_rmse_final, data = candy_train)
tidy(lasso_rmse_fit) |> 
  filter(estimate != 0) |> 
  kable()
```

## Using Parsimony as a Tie-Breaker

-   Good heuristic: **One-Standard Error Rule**
    +   Use resampling to estimate error metrics
    +   Compute standard error for error metrics
    +   Select most parsimonious model that is within one standard error of the best performance metric
    


## Selecting Best Model: Ridge

:::: columns
::: column

Best RMSE

```{r}
best_ose_rmse_ridge <- tuning_ridge_results |> 
  select_by_one_std_err(metric = "rmse", desc(penalty)) # why are we descending
best_ose_rmse_ridge |> kable(digits = 2)
```
:::

::: column
Best $R^2$
```{r}
bse_ose_rsq_ridge <- tuning_ridge_results |> 
  select_by_one_std_err(metric = "rsq", desc(penalty))
bse_ose_rsq_ridge |> kable(digits = 2)
```
:::
::::

-   Which should be use?

## Visualizing Results: Ridge

```{r}
autoplot(tuning_ridge_results)
```

## Finalize Model {.smaller}

-   RMSE estimate is less flexible and seems to be sacrificing less $R^2$ than the opposite

```{r}
ridge_rmse_final <- finalize_workflow(ridge_wf, best_ose_rmse_ridge)
ridge_rmse_fit <- fit(ridge_rmse_final, data = candy_train)
tidy(ridge_rmse_fit) |> 
  kable()
```

## Visualizing Results: LASSO

```{r}
autoplot(tuning_lasso_results)
```

## Selecting Best Model: Ridge

:::: columns
::: column

Best RMSE

```{r}
bse_ose_rmse_lasso <- tuning_lasso_results |> 
  select_by_one_std_err(metric = "rmse", desc(penalty)) # why are we descending
bse_ose_rmse_lasso |> kable(digits = 2)
```
:::

::: column
Best $R^2$
```{r}
bse_ose_rsq_lasso <- tuning_lasso_results |> 
  select_by_one_std_err(metric = "rsq", desc(penalty)) # why are we descending
bse_ose_rsq_lasso |> kable(digits = 2)
```
:::
::::

-   Which should be use?

## Finalize Model {.smaller}

-   RMSE estimate is less flexible and seems to be sacrificing less $R^2$ than the opposite

```{r}
lasso_rmse_final <- finalize_workflow(lasso_wf, bse_ose_rmse_lasso)
lasso_rmse_fit <- fit(lasso_rmse_final, data = candy_train)
tidy(lasso_rmse_fit) |> 
  filter(estimate != 0) |> 
  kable()
```

## Using the Test Set as a Tie Breaker

-   Once you've found you "best" candidate from several different classes of model, it's ok to compare on test set
-   In this case, we have our best ridge and our best lasso model
-   Main this to avoid... LOTS of comparisons on your test set

## Using the Test Set as a Tie Breaker {.smaller}

```{r}
candy_test_wpreds <- candy_test |> 
  mutate(ridge_preds = predict(ridge_rmse_fit, new_data = candy_test)$.pred,
         lasso_preds = predict(lasso_rmse_fit, new_data = candy_test)$.pred)
candy_test_wpreds |> rmse(estimate = ridge_preds, truth = winpercent)
candy_test_wpreds |> rmse(estimate = lasso_preds, truth = winpercent)
candy_test_wpreds |> rsq(estimate = ridge_preds, truth = winpercent)
candy_test_wpreds |> rsq(estimate = lasso_preds, truth = winpercent)
```

-   Ridge Wins!

## Good strategies

-   First use coarse grid with large range to find general area
    +   Then use fine grid with small range to fine tune
    +   Use iterative method to fine tune
    
# Iterative Methods (Bonus Content)

## Brief Overview

-   Basic Idea: iterative select parameter values to choose based on how previous ones have done
-   TMWR gives two examples:
    +   Bayesian search
    +   Stochastic Annealing
-   Both require an understanding of multivariate probability distributions

## Bayesian Search: GP {.smaller}

::: incremental
-   Assume performance metrics follow a Gaussian process... [dafuq]{.fragment}
-   Consider two sets of tuning parameters: $x_1 = (\lambda_1, k_1)$ and $x_2 = (\lambda_2, k_2)$
-   Let $Y_1$ and $Y_2$ be random variables representing the performance metric at these $x$'s
-   Assume $\text{Cov}(Y_1, Y_2)$ depends on how far apart $x_1$ and $x_2$
    +   Covariance should decrease the further apart $x_1$ and $x_2$ are
    +   This covariance function is typically parameterized and estimated from an initial tuning grid
:::

## Bayesian Search: Acquisition Function {.smaller}

::: incremental
-   GP process gives us predicted mean and variance of performance metrics
-   Imagine two scenario's:
    +   1: Predicted mean is slightly better than current parameter choice and variance is low
        -   Low-Risk, Low-Reward
    +   2: Predicted mean is slightly worst than current parameter choice but variance is high
        -   High-Risk, High-Reward
-   How do we balance?
-   Two competing goals: *Exploration* (got toward high variance) and *Exploitation* (go toward best mean)
-   Acquisition function: balances these two goals (several to choose from)
-   [tmwr](https://www.tmwr.org/iterative-search)
:::

## Bayesian Search: `tidymodels`

```{r}
ridge_params <- ridge_wf |> 
  extract_parameter_set_dials() |> 
  update(penalty = penalty(c(-2, 1)))

bayes_ridge <- ridge_wf |> 
  tune_bayes(
    resamples = candy_folds,
    metrics = candy_metrics, # first metrics is what's optimized (rmse in this case)
    initial = tuning_ridge_results,
    param_info = ridge_params,
    iter = 25
  )
```

## Bayesian Search: `tidymodels`

```{r}
autoplot(bayes_ridge, type = "performance")
```

## Bayesian Search: `tidymodels`

```{r}
autoplot(bayes_ridge, type = "parameters")
```

## Bayesian Search: `tidymodels`

```{r}
show_best(bayes_ridge) |> kable()
```

## Bayesian Search: `tidymodels`

```{r}
lasso_params <- lasso_wf |> 
  extract_parameter_set_dials() |> 
  update(penalty = penalty(c(-2,1)))

bayes_lasso <- lasso_wf |> 
  tune_bayes(
    resamples = candy_folds,
    metrics = candy_metrics, # first metrics is what's optimized (rmse in this case)
    initial = tuning_lasso_results,
    param_info = lasso_params,
    iter = 25
  )
```

## Bayesian Search: `tidymodels`

```{r}
autoplot(bayes_lasso, type = "performance")
```

## Bayesian Search: `tidymodels`

```{r}
autoplot(bayes_lasso, type = "parameters")
```

## Bayesian Search: `tidymodels`

```{r}
show_best(bayes_lasso) |> kable()
```

## Simulated Annealing {.smaller}

::: incremental
-   Does anyone know what annealing is in Physics/Material Science?
-   Start with initial parameter combination $x_0$(think gradient descent)
-   Embark on random walk... in each step $i$
    +   Consider small perturbation from $x_{i-1}$... let's call it $x_{i-1}^{\epsilon}$
    +   Get performance for $x_{i-1}^\epsilon$
    +   If performance is better set $x_i = x_{i-1}^\epsilon$
    +   If performance is worse set $x_i = x_{i-1}^\epsilon$ with probability $\exp(-c\times D_i\times i)$
        -   $c$ is user specified constant called cooling coefficient
        -   $D_i$ percent difference between old and new
    +   Otherwise $x_i = x_{i-1}$
-   Idea: 
    +   start hot: likely to accept suboptimal parameter combinations and move around
    +   cools over time: decrease number sub-optimal acceptances
:::

## Simulated Annealing: `tidymodels`

```{r}
library(finetune)
ridge_sim_anneal <- ridge_wf |> 
  tune_sim_anneal(
    resamples = candy_folds,
    metrics = candy_metrics, # first metric is what's optimized (rmse in this case)
    initial = tuning_ridge_results,
    param_info = ridge_params,
    iter = 50
  )
```

## Simulated Annealing `tidymodels`

```{r}
autoplot(ridge_sim_anneal, type = "performance")
```

## Simulated Annealing `tidymodels`

```{r}
autoplot(ridge_sim_anneal, type = "parameters")
```

## Simulated Annealing `tidymodels`

```{r}
show_best(ridge_sim_anneal) |> kable()
```

## Simulated Annealing: `tidymodels`

```{r}
lasso_sim_anneal <- ridge_wf |> 
  tune_sim_anneal(
    resamples = candy_folds,
    metrics = candy_metrics, # first metrics is what's optimized (rmse in this case)
    initial = tuning_lasso_results,
    param_info = lasso_params,
    iter = 50
  )
```

## Simulated Annealing `tidymodels`

```{r}
autoplot(lasso_sim_anneal, type = "performance")
```

## Simulated Annealing `tidymodels`

```{r}
autoplot(lasso_sim_anneal, type = "parameters") +
  scale_x_continuous(trans = 'log10')
```

## Simulated Annealing `tidymodels`

```{r}
show_best(lasso_sim_anneal) |> kable()
```
