---
title: 'MATH 427: More on Classification'
author: Eric Friedlander
footer: "[🔗 MAT 427 - Spring 2025 -  Schedule](https://mat427sp25.netlify.app/schedule)"
logo: "../images/logo.png"
format: 
  revealjs:
    theme: slides.scss
    multiplex: false
    transition: fade
    slide-number: false
    incremental: false 
    chalkboard: true
execute:
  freeze: auto
  echo: true
  cache: false
knitr:
  opts_chunk: 
    R.options:      
    width: 200
---


## Computational Set-Up

```{r}
library(tidyverse)
library(tidymodels)
library(rpart.plot)
library(knitr)
library(kableExtra)

tidymodels_prefer()

set.seed(427)
```

## Quick Review of Classification {.smaller}

-   We've covered many different classification methods:
    +   Logistic Regression (potentially with regularization)
    +   KNN
    +   Decision Tress
    +   Random Forests
    +   Gradient Boosted Trees
-   For each of these methods, rank them as (high/medium/low) for each of the following criteria
    +   Computational complexity to fit
    +   Computational complexity to predict
    +   Ability to handle non-linearity
    +   Interpretability
    +   Prediction accuracy
    
## Quick Review of Classification {.smaller}

|         | Logistic Regression | KNN | Decision Trees | Random Forests | Gradient Boosted Trees |
|----------|----------|----------|----------|----------|----------|
| Comp. Fit    | Low     | Low     | Low     | Medium     | High     |
| Comp. Pred.    | Low     | Depends     | Low     | Medium     | Medium     |
| Non-linearity    | Low  | Medium     | Medium     | High     | High     |
| Interp.    | High  | Low     | High     | Medium/Low     | Medium/Low     |
| Acc.    | Low  | Depends     | Low     | Medium/High     | High     |


## Data: Voter Frequency

-   [Info about data](https://github.com/fivethirtyeight/data/tree/master/non-voters)
-   Goal: Identify individuals who are unlikely to vote to help organization target "get out the vote" effort.

```{r}
voter_data <- read_csv('https://raw.githubusercontent.com/fivethirtyeight/data/master/non-voters/nonvoters_data.csv')

voter_clean <- voter_data |>
  select(-RespId, -weight, -Q1) |>
  mutate(
    educ = factor(educ, levels = c("High school or less", "Some college", "College")),
    income_cat = factor(income_cat, levels = c("Less than $40k", "$40-75k ",
                                               "$75-125k", "$125k or more")),
    voter_category = factor(voter_category, levels = c("rarely/never", "sporadic", "always"))
  ) |>
  filter(Q22 != 5 | is.na(Q22)) |>
  mutate(Q22 = as_factor(Q22),
         Q22 = if_else(is.na(Q22), "Not Asked", Q22),
         across(Q28_1:Q28_8, ~if_else(.x == -1, 0, .x)),
         across(Q28_1:Q28_8, ~ as_factor(.x)),
         across(Q28_1:Q28_8, ~if_else(is.na(.x) , "Not Asked", .x)),
         across(Q29_1:Q29_10, ~if_else(.x == -1, 0, .x)),
         across(Q29_1:Q29_10, ~ as_factor(.x)),
         across(Q29_1:Q29_8, ~if_else(is.na(.x) , "Not Asked", .x)),
        Party_ID = as_factor(case_when(
          Q31 == 1 ~ "Strong Republican",
          Q31 == 2 ~ "Republican",
          Q32 == 1  ~ "Strong Democrat",
          Q32 == 2 ~ "Democrat",
          Q33 == 1 ~ "Lean Republican",
          Q33 == 2 ~ "Lean Democrat",
          TRUE ~ "Other"
        )),
        Party_ID = factor(Party_ID, levels =c("Strong Republican", "Republican", "Lean Republican",
                                                "Other", "Lean Democrat", "Democrat", "Strong Democrat")),
        across(!ppage, ~as_factor(if_else(.x == -1, NA, .x))))
```

## Split Data

```{r}
set.seed(427)

voter_splits <- initial_split(voter_clean, prop = 0.7, strata = voter_category)
voter_train <- training(voter_splits)
voter_test <- testing(voter_splits)
```

## Problem: More than two categories

```{r}
voter_train |> 
  ggplot(aes(x = voter_category)) +
  geom_bar()
```
## Define Model: Multinomial Regression

```{r}
mn_reg_model <- multinom_reg(mixture = 1, penalty = 0.005) |> # I chose this penalty arbitrarily
  set_engine("glmnet", family = "multinomial") |> 
  set_mode("classification")
```

## Define Recipe

```{r}
mr_recipe <- recipe(voter_category ~ . , data = voter_train) |>
  step_zv(all_predictors()) |>
  step_integer(educ, income_cat, Party_ID, Q2_2:Q4_6, Q6, Q8_1:Q9_4, Q14:Q17_4,
               Q25:Q26) |>
  step_impute_median(all_numeric_predictors()) |>
  step_impute_mode(all_nominal_predictors()) |>
  step_dummy(all_nominal_predictors(), one_hot = FALSE) |> 
  step_normalize(all_numeric_predictors())
```

## Define Workflow and Fit

```{r}
mr_fit <- workflow() |>
  add_model(mn_reg_model) |>
  add_recipe(mr_recipe) |> 
  fit(voter_train)
```

## Look at Predictions

```{r}
mr_fit |> augment(new_data = voter_test) |> slice_sample(n=10) |> select(1:4) |> head() |>  kable()
```

## Confusion Matrix

```{r}
mr_fit |> augment(new_data = voter_test) |> 
  conf_mat(truth = voter_category, estimate = .pred_class) |> autoplot("heatmap")
```

## Evaluating Multiclass Models {.smaller}

-   No "Positive" and "Negative" anymore
-   Accuracy: $$\frac{\text{Correct classifications}}{\text{total observations}} = \frac{285+559+221}{1746} \approx 0.61$$
-   Most of our metrics were based on having "Positive" vs. "Negative"
    +   Precision/PPV
    +   Recall/Sensitivity
    +   Specificity
    +   NPV
    +   ROC/AUC
    
## Evaluating Multiclass Models

-   Solution 1: Apply binary metrics to each class in turn
    +   E.g. report three different Recall values
        +   Rarely: $\frac{285}{285+110+35} = \approx 0.66$
        +   Sporadic: $\frac{559}{47+559+173} = \approx 0.72$
        +   Always: $\frac{221}{21+301+221} = \approx 0.41$
-   Compute the Precision for each class:
    +   Reminder: Precision = TP/(TP+FP)
    
## Solution 1 in R

- No automatic implementation in yardstick (can hack together using `group_by` sometimes)

:::: columns
::: column
```{r}
mr_fit |> augment(new_data = voter_test) |> 
  group_by(voter_category) |> 
  recall(truth = voter_category, estimate = .pred_class) |> 
  kable()
```
:::

::: column
```{r}
mr_fit |> augment(new_data = voter_test) |> 
  group_by(voter_category) |> 
  recall(truth = voter_category, estimate = .pred_class) |> 
  kable()
```
:::
::::
    
## Evaluating Multiclass Models

-   Solution 2: Average metrics across labels
    +   *Macro-averaging* average one-versus-all metrics
        +   Recall: $\frac{0.66+0.72+0.41}{3} \approx 0.60$
    +   *Macro-weighted averaging* same but weight by class size
        +   Recall: $\frac{430\times 0.66+773\times 0.72+543\times 0.41}{1746} \approx 0.61$
    +   *Micro-averaging* compute contribution for each class, aggregates them, then computes a single metric
        +   Recall: $\frac{285+559+221}{430+779+543} \approx 0.61$
        
## Macro-Averaging in R

```{r}
mr_fit |> augment(new_data = voter_test) |> 
  recall(truth = voter_category, estimate = .pred_class, estimator = "macro") |> 
  kable()
```
        
## Macro-Weighted Averaging in R

```{r}
mr_fit |> augment(new_data = voter_test) |> 
  recall(truth = voter_category, estimate = .pred_class, estimator = "macro_weighted") |> 
  kable()
```

## Micro-Averaging in R

```{r}
mr_fit |> augment(new_data = voter_test) |> 
  recall(truth = voter_category, estimate = .pred_class, estimator = "micro") |> 
  kable()
```

## What if output is probability?

-   For binary case we used ROC curve and AUC...
-   Similar ideas apply here:
    +   One vs. all
    +   Macro Averaging
    +   NO MICRO AVERAGING!
    +   Hand and Till extension of AUC

## Plotting one-vs.-all

```{r}
mr_fit |> augment(new_data = voter_test) |> 
  roc_curve(truth = voter_category, `.pred_rarely/never`, .pred_sporadic, .pred_always) |> 
  autoplot()
```

## Macro Averaged AUC

```{r}
mr_fit |> augment(new_data = voter_test) |> 
  roc_auc(truth = voter_category, `.pred_rarely/never`, .pred_sporadic, .pred_always, 
          estimator = "macro") |> 
  kable()
```

## Macro-Weighted Averaged AUC

```{r}
mr_fit |> augment(new_data = voter_test) |> 
  roc_auc(truth = voter_category, `.pred_rarely/never`, .pred_sporadic, .pred_always, 
          estimator = "macro_weighted") |> 
  kable()
```

## Hand and Till AUC

-   [Paper](https://link.springer.com/article/10.1023/A:1010920819831)
    +   Basic Idea: Do pairwise comparison of classes and average

```{r}
mr_fit |> augment(new_data = voter_test) |> 
  roc_auc(truth = voter_category, `.pred_rarely/never`, .pred_sporadic, .pred_always) |> 
  kable()
```

## Discussion

How would heavily imbalanced classes impact each type of macro vs. micro averaging?

# Dealing with Class-Imbalance

##  Class-Imbalance

-   Class-imbalance occurs where your the classes in your response greatly differ in terms of how common they are
-   Occurs frequently:
    +   Medicine: survival/death
    +   Admission: enrollment/non-enrollment
    +   Finance: repaid loan/defaulted
    +   Tech: Clicked on ad/Didn't click
    +   Tech: Churn rate
    +   Finance: Fraud

## Data: `haberman`

Study conducted between 1958 and 1970 at the University of Chicago's Billings Hospital on the survival of patients who had undergone surgery for breast cancer.

Goal: predict whether a patient survived after undergoing surgery for breast cancer.

```{r}
haberman <- read_csv("../data/haberman.data",
                     col_names = c("Age", "OpYear", "AxNodes", "Survival"))
glimpse(haberman)
```

## Quick Clean

Study conducted between 1958 and 1970 at the University of Chicago's Billings Hospital on the survival of patients who had undergone surgery for breast cancer.

Goal: predict whether a patient *died* after undergoing surgery for breast cancer.

```{r}
haberman <- haberman |> 
  mutate(Survival = factor(if_else(Survival == 1, "Survived", "Died"),
                           levels = c("Died", "Survived")))
glimpse(haberman)
```

## Split Data

```{r}
set.seed(427)
hab_splits <- initial_split(haberman, prop = 0.75, strata = Survival)
hab_train <- training(hab_splits)
hab_test <- testing(hab_splits)
```

## Visualizing Response

```{r}
hab_train |> 
  ggplot(aes(y = Survival)) +
  geom_bar()
```

## Fitting Model

```{r}
lr_fit <- logistic_reg() |> 
  set_engine("glm") |> 
  fit(Survival ~ . , data = hab_train)
```

## Confusion Matrix

```{r}
lr_fit |> augment(new_data = hab_test) |> 
  conf_mat(truth = Survival, estimate = .pred_class) |> autoplot("heatmap")
```

## Performance Metrics

```{r}
hab_metrics <- metric_set(accuracy, precision, recall)

lr_fit |> augment(new_data = hab_test) |> 
  roc_auc(truth = Survival, .pred_Died) |> kable()

lr_fit |> augment(new_data = hab_test) |> 
  hab_metrics(truth = Survival, estimate = .pred_class) |> kable()
```

## Recall is BAD!

-   Since there are so few deaths, model always predicts a low probability of death
-   Idea: just because you you have a HIGHER probability of death doesn't mean have a HIGH probability of death

## What do we do?

-   Depends on what your goal is...
-   Ask yourself: What is most important to my problem?
    +   Accurate probabilities?
    +   Effective Separation?
    +   Effective identification of positives?
    +   Low false-positive rate?
-   Discussion: Let's think of scenarios where each one of these is the most important.

