---
title: 'MATH 427: More More on Classification'
author: Eric Friedlander
footer: "[🔗 MAT 427 - Spring 2025 -  Schedule](https://mat427sp25.netlify.app/schedule)"
logo: "../images/logo.png"
format: 
  revealjs:
    theme: slides.scss
    multiplex: false
    transition: fade
    slide-number: false
    incremental: false 
    chalkboard: true
execute:
  freeze: auto
  echo: true
  cache: false
knitr:
  opts_chunk: 
    R.options:      
    width: 200
---


## Computational Set-Up

```{r}
library(tidyverse)
library(tidymodels)
library(knitr)
library(kableExtra)

tidymodels_prefer()

set.seed(427)
```



## Data: Voter Frequency

-   [Info about data](https://github.com/fivethirtyeight/data/tree/master/non-voters)
-   Goal: Identify individuals who are unlikely to vote to help organization target "get out the vote" effort.

```{r}
voter_data <- read_csv('https://raw.githubusercontent.com/fivethirtyeight/data/master/non-voters/nonvoters_data.csv')

voter_clean <- voter_data |>
  select(-RespId, -weight, -Q1) |>
  mutate(
    educ = factor(educ, levels = c("High school or less", "Some college", "College")),
    income_cat = factor(income_cat, levels = c("Less than $40k", "$40-75k ",
                                               "$75-125k", "$125k or more")),
    voter_category = factor(voter_category, levels = c("rarely/never", "sporadic", "always"))
  ) |>
  filter(Q22 != 5 | is.na(Q22)) |>
  mutate(Q22 = as_factor(Q22),
         Q22 = if_else(is.na(Q22), "Not Asked", Q22),
         across(Q28_1:Q28_8, ~if_else(.x == -1, 0, .x)),
         across(Q28_1:Q28_8, ~ as_factor(.x)),
         across(Q28_1:Q28_8, ~if_else(is.na(.x) , "Not Asked", .x)),
         across(Q29_1:Q29_10, ~if_else(.x == -1, 0, .x)),
         across(Q29_1:Q29_10, ~ as_factor(.x)),
         across(Q29_1:Q29_8, ~if_else(is.na(.x) , "Not Asked", .x)),
        Party_ID = as_factor(case_when(
          Q31 == 1 ~ "Strong Republican",
          Q31 == 2 ~ "Republican",
          Q32 == 1  ~ "Strong Democrat",
          Q32 == 2 ~ "Democrat",
          Q33 == 1 ~ "Lean Republican",
          Q33 == 2 ~ "Lean Democrat",
          TRUE ~ "Other"
        )),
        Party_ID = factor(Party_ID, levels =c("Strong Republican", "Republican", "Lean Republican",
                                                "Other", "Lean Democrat", "Democrat", "Strong Democrat")),
        across(!ppage, ~as_factor(if_else(.x == -1, NA, .x))))
```

## Split Data

```{r}
set.seed(427)

voter_splits <- initial_split(voter_clean, prop = 0.7, strata = voter_category)
voter_train <- training(voter_splits)
voter_test <- testing(voter_splits)
```

## Problem: More than two categories

```{r}
voter_train |> 
  ggplot(aes(x = voter_category)) +
  geom_bar()
```

## Define Model: Multinomial Regression

```{r}
mn_reg_model <- multinom_reg(mixture = 1, penalty = 0.005) |> # I chose this penalty arbitrarily
  set_engine("glmnet", family = "multinomial") |> 
  set_mode("classification")
```

## Define Recipe

```{r}
mr_recipe <- recipe(voter_category ~ . , data = voter_train) |>
  step_zv(all_predictors()) |>
  step_integer(educ, income_cat, Party_ID, Q2_2:Q4_6, Q6, Q8_1:Q9_4, Q14:Q17_4,
               Q25:Q26) |>
  step_impute_median(all_numeric_predictors()) |>
  step_impute_mode(all_nominal_predictors()) |>
  step_dummy(all_nominal_predictors(), one_hot = FALSE) |> 
  step_normalize(all_numeric_predictors())
```

## Define Workflow and Fit

```{r}
mr_fit <- workflow() |>
  add_model(mn_reg_model) |>
  add_recipe(mr_recipe) |> 
  fit(voter_train)
```

## Look at Predictions

```{r}
mr_fit |> augment(new_data = voter_test) |> slice_sample(n=10) |> select(1:4) |> head() |>  kable()
```

## Confusion Matrix

```{r}
mr_fit |> augment(new_data = voter_test) |> 
  conf_mat(truth = voter_category, estimate = .pred_class) |> autoplot("heatmap")
```

## Last Time {.smaller}

-   No "Positive" and "Negative" anymore
-   Most of our metrics were based on having "Positive" vs. "Negative"
-   Solution 1: 1-vs-all metrics
    
## Evaluating Multiclass Models

-   Solution 2: Average metrics across labels
    +   *Macro-averaging* average one-versus-all metrics
        +   Recall: $\frac{0.66+0.72+0.41}{3} \approx 0.60$
    +   *Macro-weighted averaging* same but weight by class size
        +   Recall: $\frac{430\times 0.66+773\times 0.72+543\times 0.41}{1746} \approx 0.61$
    +   *Micro-averaging* compute contribution for each class, aggregates them, then computes a single metric
        +   Recall: $\frac{285+559+221}{430+779+543} \approx 0.61$
        
## Questions From Last Time

-   Multinomial Logistic regression:
    +   Each class $k$: $$\log \frac{\text{Prob. class } k}{\text{Prob. class } K} = \beta_{0k} + \beta_{1k}X_1 + \cdots + \beta_{pk}X_p$$
-   Shuba: Micro-averaged recall is the same as accuracy... CORRECT!
    +   Same is true of precision AND $F_1$
    +   Useful if you have a "multi-label" classification problem (not covered in this class)
    
## Rabin's Question {.smaller}

-   "What does a medium sized data set mean?"
-   Idea behind ML: identify patterns in data and use them to make predictions
-   As data gets "bigger":
    +   Patterns become clearer
    +   Computational complexity increases
-   Informal definitions:
    +   Small data: not enough data to fully represent patterns
    +   Big data: all the info is there but special approaches need to be taken to handle all the data
    
## Thinking about small data

-   Patterns not fully represented in data $\Rightarrow$ restrict the set of possible patterns and give model less flexibility and freedom
    +   Logistic regression (probably with regularization)
    +   Support vector machines (we haven't talked about these yet)
    +   To a lesser extent: Decision Trees
    +   NOT KNN

## Thinking about big data {.smaller}

-   Patterns are definitely there but size introduces computational problems
    +   Data set can't fit in memory
        +   Solution 1: Use a high-memory HPC cluster node
        +   Solution 2: Modify algorithms to use parts of data instead of full data set (e.g. Stochastic gradient descent)
    +   Algorithm scales with size of data and will take too long to run/fit
        +   Solution 1: Run in parallel if possible using HPC cluster
        +   Solution 2: Develop faster algorithms
            +   Implement in a compiled language like C
            +   Develop (faster) approximate solution
-   Curse of dimensionalality
    +   If $p$ is too-big $\Rightarrow$ too much space and things are too far apart $\Rightarrow$ similar impact to small data but without computational benefit
-   Note: big $n$ vs. big $p$ can present different issues
            
## Medium Data

-   Data that's big enough to have (most) of the information you need but not so big that it presents computational issues
-   KNN sweet spot
    +   Enough data the "nearest-neighbors" are actually "near"
    +   Not so much data that it takes forever to make predictions
    
## Exploring with App

-   [App 1](https://efriedlander.shinyapps.io/ClassificationMetrics/)
-   [App 2](https://efriedlander.shinyapps.io/BVTappClassification/)
        
## Macro-Averaging in R

```{r}
voter_metrics <- metric_set(accuracy, precision, recall)

mr_fit |> augment(new_data = voter_test) |> 
  voter_metrics(truth = voter_category, estimate = .pred_class, estimator = "macro") |> 
  kable()
```
        
## Macro-Weighted Averaging in R

```{r}
mr_fit |> augment(new_data = voter_test) |> 
  voter_metrics(truth = voter_category, estimate = .pred_class, estimator = "macro_weighted") |> 
  kable()
```

## Micro-Averaging in R

```{r}
mr_fit |> augment(new_data = voter_test) |> 
  voter_metrics(truth = voter_category, estimate = .pred_class, estimator = "micro") |> 
  kable()
```

## What if output is probability?

-   For binary case we used ROC curve and AUC...
-   Similar ideas apply here:
    +   One vs. all
    +   Macro Averaging
    +   NO MICRO AVERAGING!
    +   Hand and Till extension of AUC

## Plotting one-vs.-all

```{r}
mr_fit |> augment(new_data = voter_test) |> 
  roc_curve(truth = voter_category, `.pred_rarely/never`, .pred_sporadic, .pred_always) |> 
  autoplot()
```

## Macro Averaged AUC

```{r}
mr_fit |> augment(new_data = voter_test) |> 
  roc_auc(truth = voter_category, `.pred_rarely/never`, .pred_sporadic, .pred_always, 
          estimator = "macro") |> 
  kable()
```

## Macro-Weighted Averaged AUC

```{r}
mr_fit |> augment(new_data = voter_test) |> 
  roc_auc(truth = voter_category, `.pred_rarely/never`, .pred_sporadic, .pred_always, 
          estimator = "macro_weighted") |> 
  kable()
```

## Hand and Till AUC {.smaller}

-   Idea behind traditional AUC: "How well are my classes being separated?"
-   Hand and Till: Extend this idea to multiple-classes
-   [Paper](https://link.springer.com/article/10.1023/A:1010920819831)
    +   Basic Idea: Do pairwise comparison of classes and average

```{r}
mr_fit |> augment(new_data = voter_test) |> 
  roc_auc(truth = voter_category, `.pred_rarely/never`, .pred_sporadic, .pred_always) |> 
  kable()
```

## Discussion

-   Why do you compute averages/means?

. . .

-   How would heavily imbalanced classes impact each type of averaging?
    +   Which type(s) of averaging weight(s) each class equally regardless of balance?
    +   Which type(s) of averaging favor(s) larger classes?
    +   Does this imply that one is better than the others?
    
## Exploring with App {.smaller}

-   [App](https://efriedlander.shinyapps.io/ClassificationMetrics/)
    +   Break into groups
    +   Investigate how your performance metrics change between balanced data and unbalanced data
    +   Additional Considerations:
        +   Impact of boundaries/models?
        +   Impact of sample size?
        +   Impact of noise level?
    +   Please write down observations so we can discuss
    
