---
title: 'MATH 427: Principal Component Analysis (PCA)'
author: Eric Friedlander
footer: "[🔗 MAT 427 - Spring 2025 -  Schedule](https://mat427sp25.netlify.app/schedule)"
logo: "../images/logo.png"
format: 
  revealjs:
    theme: slides.scss
    multiplex: false
    transition: fade
    slide-number: false
    incremental: false 
    chalkboard: true
execute:
  freeze: auto
  echo: true
  cache: true
knitr:
  opts_chunk: 
    R.options:      
    width: 200
---

## Computational Set-Up

```{r}
library(tidyverse)
library(tidymodels)
library(knitr)
library(kableExtra)

tidymodels_prefer()

set.seed(427)
```

## Data: `mnist`

-   MNIST Database: Modified National Institute of Standards and Technology Database
-   Large database of handwritten digits
    +   60,000 training images
    +   10,000 test images
-   Each image:
    +   28x28 black and white pixels
    +   $28\times 28\times 1 = 784$
    
## Loading data

```{r}
#| cache: TRUE

library(dslabs)
mnist <- read_mnist()
mnist_train <- mnist$train$images
mnist_train |> head() |> kable()
```

## Digits

```{r}
#| code-fold: TRUE

image(x = 1:28, y = 1:28,
      z = matrix(mnist_train[1,], nrow = 28, byrow=FALSE)[,28:1],
      col=gray((0:255)/255))
```

## Digits

```{r}
#| code-fold: TRUE

image(x = 1:28, y = 1:28,
      z = matrix(mnist_train[2,], nrow = 28, byrow=FALSE)[,28:1],
      col=gray((0:255)/255))
```

## Digits

```{r}
#| code-fold: TRUE

image(x = 1:28, y = 1:28,
      z = matrix(mnist_train[3,], nrow = 28, byrow=FALSE)[,28:1],
      col=gray((0:255)/255))
```

## Digits

```{r}
#| code-fold: TRUE

image(x = 1:28, y = 1:28,
      z = matrix(mnist_train[4,], nrow = 28, byrow=FALSE)[,28:1],
      col=gray((0:255)/255))
```
    
## Unsupervised Learning & Dimensionality Reduction

-   **Unsupervised Learning**: ML for unlabeled data (i.e. no response variables)
    +   Goal: Uncover patterns/structure within data
    +   Tasks:
        -   **Clustering**: finding sub-groups within our data
        -   **Dimensionality Reduction**: reducing the number of columns in our data set... [why?]{.fragment}
        
## Dimensionality Reduction

-   Goal phrasing 1: Reduce the number of columns, while losing as little information as possible
-   Goal phrasing 2: Extract lower-dimensional structure from our data

## Last Time

-   Vectors and Projects

# Principal Component Analysis (PCA)

## PCA Vocabulary

-   **Principal Component (PC1)**: direction in $p$-dimensional space (e.g. $\langle 1, 1, 2\rangle$)
-   **Scores**: our new variables (e.g. $(-0.56\times 1 + -0.996\times 1 + -1.56\times 2)/6 = -0.778$)
-   **Loadings**: For direction above'
    +   Loading on $x$ is 1
    +   Loading on $y$ is 1
    +   Loading on $z$ is 2

## Recall: Variance

::: incremental
-   What is variance?
-   Intuitively: what does variance measure?
-   Variance: $\frac{1}{n-1}\sum_{i=1}^n(x_i - \bar{x})^2$
    +   Average of the squared distance from zero of each observation
:::

## Idea behind PCA

-   Select first PC so variance of scores is the maximum
-   Iteratively:
    +   Select next PC so variance of scores is maximize AND new PC is *orthogonal* to all other PCs

. . .

-   What does orthogonal mean?

## Easy Example

-   Exercise: What should the first and second PCs be?

```{r}
#| echo: false
#| code-fold: true

set.seed(123)
n = 1000
easy_ex <- tibble(x = rnorm(n), y = rnorm(n, sd = 0.25))

ggplot(easy_ex, aes(x=x, y=y)) +
  geom_point() +
  xlim(-4, 4)+
  ylim(-4,4)
```

## How much variance is explained by each of the PC's?

```{r}
var_exp <- easy_ex |> 
  mutate(PC1 = x,
         PC2 = y) |> 
  summarize(var1 = var(PC1), 
            var2 = var(PC2))

var_exp |> kable()
```

## What proportion of variance is explained by each of the PC's?

```{r}
var_exp |> 
  pivot_longer(everything()) |> 
  mutate(proportion = value/sum(value)) |> 
  kable()
```

-   93% of our variance (information) is contained in our first PC

## Harder Example

-   Exercise: What should the first and second PCs be?

```{r}
#| echo: false
#| code-fold: true

set.seed(123)
n = 1000
harder_ex <- tibble(x = rnorm(n), y = x + rnorm(n, sd = 0.25))

ggplot(harder_ex, aes(x=x, y=y)) +
  geom_point() +
  xlim(-4, 4)+
  ylim(-4,4)
```

## How much variance is explained by each of the PC's?

```{r}
var_exp <- harder_ex |> 
  mutate(PC1 = (x + y)/2,
         PC2 = (x-y)/2) |> 
  summarize(var1 = var(PC1), 
            var2 = var(PC2))

var_exp |> kable()
```

## Question?

::: incremental
-   Will our PCs depend on the scale of our data?
-   What does this mean we should do?
:::

## How is PCA actually accomplished

-   Compute **Variance-Covariance Matrix**
-   Compute **Eigen-decomposition of variance-covariance matrix**
    +   Eigen-vectors = Principal components
    +   Eigen-values = Variance explained by corresponding component
    
# PCA in R

## Recipe

-   [Documentation for `step_normalize`](https://recipes.tidymodels.org/reference/step_pca.html)

```{r}
pca_recipe <- recipe(~ ., data = mnist_train) |> 
  step_normalize(all_numeric_predictors()) |> 
  step_pca(all_numeric_predictors(), num_comp = 20)
```

## Prepping and Baking `recipe`s

-   If we want to apply a recipe to a data set *outside of a workflow*:
    +   First, `prep` using training set: this estimates any necessary quantities
    +   Second, `bake` this applies the recipe to a new data set (in our case the new data set in the same data set)
    
## Applying recipe to training data

```{r}
prepped_pca <- pca_recipe |> prep(mnist_train)
mnist_pca <- prepped_pca |> bake(new_data = mnist_train)
```

## Using the results

-   Three things of interest
    +   New scores: this is our new data... use for plotting and analysis
    +   PCs: these are our new directions... by looking at loadings we can interpret structure
    +   Eigenvalues: this will tell us how much information is lost
    
## How many PCs Should We Use?

:::: columns
::: column
```{r}
prepped_pca |> 
  tidy(2, type = "variance") |> 
  filter(terms == "variance") |> 
  kable()
```
:::

::: column
```{r}
prepped_pca |> 
  tidy(2, type = "variance") |> 
  filter(terms == "percent variance") |> 
  kable()
```
:::
::::

## Scree Plot

-   Percent variance against number of components

```{r}
#| code-fold: true

prepped_pca |> 
  tidy(2, type = "variance") |> 
  filter(terms == "percent variance") |> 
  ggplot(aes(x = component, y = value)) +
  geom_line() +
  labs(
    x = "PCAs",
    y = "% of variance",
    title = "Scree plot"
  )
```

## Scree Plot

-   Percent variance against number of components

```{r}
#| code-fold: true

prepped_pca |> 
  tidy(2, type = "variance") |> 
  filter(terms == "percent variance", component < 40) |>
  ggplot(aes(x = component, y = value)) +
  geom_line() +
  labs(
    x = "PCAs",
    y = "% of variance",
    title = "Scree plot"
  )
```

## Heuristic

-   Set number of components to be where you see a "big drop"
-   Choose number of components where plot starts to level off
    
## New Scores

```{r}
mnist_pca |> head() |> kable()
```

## Visualizing Scores

```{r}
#| code-fold: true

library(GGally)

mnist_pca |> ggpairs(columns = 1: 5, aes(color = as_factor(mnist$train$labels)),
                     lower = list(continuous = wrap("points", alpha = 0.1,    size=0.1)))
```

## Visualizing PCs

```{r}
#| code-fold: true

prepped_pca |> tidy(2) |> 
  filter(component == "PC1") |> 
  ggplot(aes(x = terms, y = value)) +
  geom_col() +
  theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank())
```

## Visualizing PC1

```{r}
#| code-fold: true

pc_matrix <- prepped_pca |> tidy(2) |> 
  filter(component == "PC1") |> 
  pull(value) |> 
  matrix(nrow = 28, byrow=FALSE)

image(x = 1:28, y = 1:28,
      z = pc_matrix[,28:1],
      col=gray((0:255)/255))
```

## Visualizing PC2

```{r}
#| code-fold: true

pc_matrix <- prepped_pca |> tidy(2) |> 
  filter(component == "PC2") |> 
  pull(value) |> 
  matrix(nrow = 28, byrow=FALSE)

image(x = 1:28, y = 1:28,
      z = pc_matrix[,28:1],
      col=gray((0:255)/255))
```

## Visualizing PCs

```{r}
#| code-fold: true

pc_matrix <- prepped_pca |> tidy(2) |> 
  filter(component == "PC3") |> 
  pull(value) |> 
  matrix(nrow = 28, byrow=FALSE)

image(x = 1:28, y = 1:28,
      z = pc_matrix[,28:1],
      col=gray((0:255)/255))
```

## Twos and Sevens

```{r}
mnist_train_27 <- mnist_train[mnist$train$labels %in% c(2, 7),]
prepped_pca27 <- pca_recipe |> prep(mnist_train_27)
mnist27_pca <- prepped_pca27 |> bake(new_data = mnist_train_27)
```

## Visualizing PCs

```{r}
#| code-fold: true

pc_matrix <- prepped_pca27 |> tidy(2) |> 
  filter(component == "PC1") |> 
  pull(value) |> 
  matrix(nrow = 28, byrow=FALSE)

image(x = 1:28, y = 1:28,
      z = pc_matrix[,28:1],
      col=gray((0:255)/255))
```

## Visualizing PCs

```{r}
#| code-fold: true

pc_matrix <- prepped_pca27 |> tidy(2) |> 
  filter(component == "PC2") |> 
  pull(value) |> 
  matrix(nrow = 28, byrow=FALSE)

image(x = 1:28, y = 1:28,
      z = pc_matrix[,28:1],
      col=gray((0:255)/255))
```

## Visualizing PCs

```{r}
#| code-fold: true

pc_matrix <- prepped_pca27 |> tidy(2) |> 
  filter(component == "PC3") |> 
  pull(value) |> 
  matrix(nrow = 28, byrow=FALSE)

image(x = 1:28, y = 1:28,
      z = pc_matrix[,28:1],
      col=gray((0:255)/255))
```