---
title: 'MAT-427: Data Splitting + KNN'
author: Eric Friedlander
footer: "[🔗 MAT 427 - Spring 2025 -  Schedule](https://mat427sp25.netlify.app/schedule)"
logo: "../images/logo.png"
format: 
  revealjs:
    theme: slides.scss
    multiplex: false
    transition: fade
    slide-number: false
    incremental: false 
    chalkboard: true
execute:
  freeze: auto
  echo: true
knitr:
  opts_chunk: 
    R.options:      
    width: 200
---


## Computational Setup

```{r setup}
library(tidyverse)
library(tidymodels)
library(knitr)
library(readODS)
library(modeldata) # contains ames dataset

tidymodels_prefer()

mlr_model <- linear_reg() |> 
  set_engine("lm")
```

## Comparing Models: Data Splitting {.smaller}

- Split `ames` data set into two parts
  + Training set: randomly selected proportion $p$ (typically 50-90%) of data used for fitting model
  + Test set: randomly selected proportion $1-p$ of data used for estimating prediction error
- If comparing A LOT of models, split into *three* parts to prevent **information leakage**
  + Training set: randomly selected proportion $p$ (typically 50-90%) of data used for fitting model
  + Validation set: randomly selected proportion $q$ (typically 20-30%) of data used to choosing tuning parameters
  + Test set: randomly selected proportion $1-p-q$ of data used for estimating prediction error
- Idea: use data your model hasn't seen to get more accurate estimate of error and prevent overfitting

## Comparing Models: Data Splitting with `tidymodels` {.smaller}

```{r}
set.seed(427) # Why?

ames_split <- initial_split(ames, prop = 0.70, strata = Sale_Price) # initialize 70/30
ames_split

ames_train <- training(ames_split) # get training data
ames_test <- testing(ames_split) # get test data
```

- `strata` not necessary but good practice
  + `strata` will use *stratified sampling* on the variable you specify (very little downside) 

## Linear Regression: Comparing Models {.smaller}

- Let's create three models with `Sale_Price` as the response:
  + **fit1**: a linear regression model with `Bedroom_AbvGr`  as the only predictor
  + **fit2**: a linear regression model with `Gr_Liv_Area` as the only predictor
  + **fit3** (similar to model in previous slides): a multiple regression model with `Gr_Liv_Area` and `Bedroom_AbvGr` as predictors
  + **fit4**: super flexible model which fits a 10th degree polynomial to `Gr_Liv_Area` and a 2nd degree polynomial to `Bedroom_AbvGr`

```{r}
fit1 <- mlr_model |> fit(Sale_Price ~ Bedroom_AbvGr, data = ames_train) # Use only training set
fit2 <- mlr_model |> fit(Sale_Price ~ Gr_Liv_Area, data = ames_train)
fit3 <- mlr_model |> fit(Sale_Price ~ Gr_Liv_Area + Bedroom_AbvGr, data = ames_train)
fit4 <- mlr_model |> fit(Sale_Price ~ poly(Gr_Liv_Area, degree = 10) + poly(Bedroom_AbvGr, degree = 2), data = ames_train)
```

## Computing MSE {.smaller}

```{r}
# Fit 1
fit1_train_mse <- mean((ames_train$Sale_Price - predict(fit1, new_data = ames_train)$.pred)^2)
fit1_test_mse <- mean((ames_test$Sale_Price - predict(fit1, new_data = ames_test)$.pred)^2)

# Fit 2
fit2_train_mse <- mean((ames_train$Sale_Price - predict(fit2, new_data = ames_train)$.pred)^2)
fit2_test_mse <- mean((ames_test$Sale_Price - predict(fit2, new_data = ames_test)$.pred)^2)

# Fit 
fit3_train_mse <- mean((ames_train$Sale_Price - predict(fit3, , new_data = ames_train)$.pred)^2)
fit3_test_mse <- mean((ames_test$Sale_Price - predict(fit3, new_data = ames_test)$.pred)^2)

# Fit 
fit4_train_mse <- mean((ames_train$Sale_Price - predict(fit4, , new_data = ames_train)$.pred)^2)
fit4_test_mse <- mean((ames_test$Sale_Price - predict(fit4, new_data = ames_test)$.pred)^2)
```

## [Question]{style="color:blue"}

Without looking at the numbers

1. Do we know which of the following is the smallest: `fit1_train_mse`, `fit2_train_mse`, `fit3_train_mse`, `fit4_train_mse`? [Yes, `fit4_train_mse`]{.fragment .fade-in}
2. Do we know which of the following is the smallest: `fit1_test_mse`, `fit2_test_mse`, `fit3_test_mse`, `fit4_test_mse`? [No]{.fragment .fade-in}

## Choosing a Model {.smaller}

::::{.columns}
:::{.column}
```{r}
# Training Errors
c(fit1_train_mse, fit2_train_mse, 
  fit3_train_mse, fit4_train_mse)
which.min(c(fit1_train_mse, fit2_train_mse, 
            fit3_train_mse, fit4_train_mse))

# test Errors
c(fit1_test_mse, fit2_test_mse, 
  fit3_test_mse, fit4_test_mse)
which.min(c(fit1_test_mse, fit2_test_mse, 
            fit3_test_mse, fit4_test_mse))
```
:::
:::{.column}
- `fit4` has the lowest training MSE (to be expected)
- `fit3` has the lowest test MSE
  + We would choose `fit3`
- Anything else interesting we see?
:::
::::

# K-Nearest Neighbors

## Regression: Conditional Averaging {.smaller}

**Restaurant Outlets Profit dataset**

```{r}
#| echo: FALSE
outlets <- readRDS("../data/outlets.rds")   # load dataset

ggplot(data = outlets, aes(x = population, y = profit)) +
  geom_point() +   # create scatterplot
  geom_smooth(method = "lm", se = FALSE)   # add the SLR line
```


What is a good value of $\hat{f}(x)$ (expected profit), say at $x=6$?

A possible choice is the **average of the observed responses** at $x=6$. But we may not observe responses for certain $x$ values.


## K-Nearest Neighbors (KNN) Regression  {.smaller}

- Non-parametric approach
- Formally: Given a value for $K$ and a test data point $x_0$,
$$\hat{f}(x_0)=\dfrac{1}{K} \sum_{x_i \in \mathcal{N}_0} y_i=\text{Average} \ \left(y_i \ \text{for all} \ i:\ x_i \in \mathcal{N}_0\right) $$
where $\mathcal{N}_0$ is the set of the $K$ training observations closest to $x_0$.
- Informally, average together the $K$ "closest" observations in your training set
- "Closeness": usually use the **Euclidean metric** to measure distance
- Euclidean distance between $\mathbf{X}_i=(x_{i1}, x_{i2}, \ldots, x_{ip})$ and $\mathbf{x}_j=(x_{j1}, x_{j2}, \ldots, x_{jp})$:
$$||\mathbf{x}_i-\mathbf{x}_j||_2 = \sqrt{(x_{i1}-x_{j1})^2 + (x_{i2}-x_{j2})^2 + \ldots + (x_{ip}-x_{jp    })^2}$$

## [KNN Regression (single predictor): Fit]{.r-fit-text} {.smaller}

::::{.columns}
:::{.column}
**$K=1$**
```{r}
knnfit1 <- nearest_neighbor(neighbors = 1) |> 
  set_engine("kknn") |> 
  set_mode("regression") |> 
  fit(profit ~ population, data = outlets)   # 1-nn regression
predict(knnfit1, new_data = tibble(population = 6)) |> kable()  # 1-nn prediction
```

```{r}
#| echo: FALSE
outlets |>
  mutate(dist = abs(population - 6),
         is_knn = if_else(dist <= min(dist), TRUE, FALSE)) |>
  ggplot(aes(x = population, y = profit, color = is_knn)) +
  geom_point(size=3) +
  geom_vline(xintercept = 6) +
  theme(text = element_text(size = 20))
```


:::
:::{.column}
**$K=5$**
```{r}
knnfit5 <- nearest_neighbor(neighbors = 5) |> 
  set_engine("kknn") |> 
  set_mode("regression") |> 
  fit(profit ~ population, data = outlets)   # 1-nn regression
predict(knnfit5, new_data = tibble(population = 6)) |> kable()  # 1-nn prediction
```

```{r}
#| echo: FALSE
outlets |>
  mutate(dist = abs(population - 6),
         is_knn = if_else(dist <= nth(dist, n=-5), TRUE, FALSE)) |>
  ggplot(aes(x = population, y = profit, color = is_knn)) +
  geom_point(size = 3) +
  geom_vline(xintercept = 6) +
  theme(text = element_text(size = 20))
```
:::
::::

## Regression Methods: Comparison

```{r}
#| echo: FALSE

slrfit <- linear_reg() |> 
  set_engine("lm") |> 
  fit(profit ~ population, data = outlets)   # fit the SLR model
pop_seq <- tibble(population = seq(min(outlets$population, na.rm = TRUE), max(outlets$population, na.rm = TRUE), 0.01))

# obtain predictions for all training data points
knn_1 <- predict(knnfit1, new_data = pop_seq)
knn_5 <- predict(knnfit5, new_data = pop_seq)
p <-  predict(slrfit, new_data = pop_seq)

# column bind original data with predicted values
predictions <- pop_seq |> 
  bind_cols(linear = p$.pred, knn_1 = knn_1$.pred, knn_5 = knn_5$.pred) |>
  pivot_longer(cols = !population, names_to = "Method", values_to = "profit")


# plot the three models
ggplot(data = outlets, aes(x = population, y = profit)) +
  geom_point() +
  geom_line(data = predictions, aes(x = population, y = profit, color = Method, linetype = Method), linewidth = 1)
```


## <span style="color:blue">Question!!!</span>

As $K$ in KNN regression increases:

- the flexibility of the fit ([decreases]{.fragment .highlight-red} /increases)
- the bias of the fit (decreases/[increases]{.fragment .highlight-red} )
- the variance of the fit ([decreases]{.fragment .highlight-red}/increases)


## [K-Nearest Neighbors Regression (multiple predictors)]{.r-fit-text} {.smaller}

- Let's look at the `ames` data
```{r}
#| echo: TRUE

ames |>
  select(Sale_Price, Gr_Liv_Area, Bedroom_AbvGr) |>
  head() |> 
  kable()
```
:::{.fragment}
:::{.incremental}
- Should 1 square foot count the same as 1 bedroom?
- Need to **center and scale** (freq. just say scale)
  + subtract mean from each predictor
  + divide by standard deviation of each predictor
  + compares apples-to-apples
:::
:::

## Scaling in R

```{r}
# scale predictors
ames_scaled <- tibble(size_scaled = scale(ames$Gr_Liv_Area),
                                  num_bedrooms_scaled = scale(ames$Bedroom_AbvGr),
                                  price = ames$Sale_Price)

head(ames_scaled) |> kable()  # first six observations
```

## Question...

:::{.incremental}
-   What about the training and test sets?
-   Need to scale BOTH sets based on the mean and standard deviation of the training set...
-   Discussion: Why?
-   Discussion: Why don't I need to center and scale `Sale_Price`?
:::

## Scaling Revisited

```{r}
ames_train_scaled <- tibble(size_scaled = scale(ames_train$Gr_Liv_Area),
                                  num_bedrooms_scaled = scale(ames_train$Bedroom_AbvGr),
                                  price = ames_train$Sale_Price)

ames_test_scaled <- tibble(size_scaled = (ames_test$Gr_Liv_Area - mean(ames_train$Gr_Liv_Area)/sd(ames_train$Gr_Liv_Area)),
                                  num_bedrooms_scaled = (ames_test$Bedroom_AbvGr - mean(ames_train$Bedroom_AbvGr))/sd(ames_train$Bedroom_AbvGr),
                                  price = ames_test$Sale_Price)
```

-   Next time: using `recipe`'s in `tidymodels` to simplify this process

## [K-Nearest Neighbors Regression (multiple predictors)]{.r-fit-text} {.smaller}


```{r}
knnfit10 <- nearest_neighbor(neighbors = 10) |>   # 10-nn regression
  set_engine("kknn") |> 
  set_mode("regression") |> 
  fit(price ~ size_scaled + num_bedrooms_scaled, data = ames_train_scaled)
```

- Test Point: `Gr_Liv_area` = 2000 square feet, and `Bedroom_AbvGr` = 3, then

```{r}
# obtain 10-nn prediction

predict(knnfit10, new_data = tibble(size_scaled = (2000 - mean(ames_train$Gr_Liv_Area))/sd(ames_train$Gr_Liv_Area),
                                     num_bedrooms_scaled = (3 - mean(ames_train$Bedroom_AbvGr))/sd(ames_train$Bedroom_AbvGr)))
```

