---
title: 'MATH 427: The Big Picture'
author: Eric Friedlander
footer: "[🔗 MAT 427 - Spring 2025 -  Schedule](https://mat427sp25.netlify.app/schedule)"
logo: "../images/logo.png"
format: 
  revealjs:
    theme: slides.scss
    multiplex: false
    transition: fade
    slide-number: false
    incremental: false 
    chalkboard: true
execute:
  freeze: auto
  echo: true
knitr:
  opts_chunk: 
    R.options:      
    width: 200
---

```{r setup}
#| include: false

library(tidyverse)
library(ISLR2)
library(kableExtra)
library(countdown)
```


# Having a successful semester in MAT 427

## Five tips for success

1.  Complete all the preparation work (readings and videos) before class.

2.  Ask questions, come to office hours and help session.

3.  Do the homework; get started on homework early when possible.

4.  Don't procrastinate and don't let a week pass by with lingering questions.

5.  Stay up-to-date on announcements on Canvas and sent via email.

## Emails for help

If you email me about an error please include a screenshot of the error and the code causing the error.

# Questions?

Raise your hand or email me.

# Machine Learning

## What is Machine Learning?

-   Machine Learning is the study of tools/techniques for extracting
    information and making predictions from complex datasets
-   The name machine learning was coined in 1959 by Arthur Samuel
    -   "Field of study that gives computers the ability to learn
        without being explicitly programmed"

## What is Machine Learning?

Tom M. Mitchell (1998):

> A computer program is said to learn from experience **E** with respect
> to some class of tasks **T** and performance measure **P** if its
> performance at tasks in **T**, as measured by **P**, improves with
> experience **E**.

## What is Machine Learning?

![MNIST handwritten digits (from ISLR, James et
al.)](images/01/MNIST-0000000001-2e09631a_09liOmx.jpg)

## [Question!!!]{style="color:blue"}

Suppose your email program watches which emails you do or do not mark as
spam, and based on that learns how to better filter spam. According to
Tom Mitchell's definition, which of the following is the task **T**,
experience **E**, and performance measure **P** in this setting?

-   [**P**]{.fragment .fade-in fragment-index="1"} The number (or
    fraction) of emails correctly classified as spam/ham (not spam)
-   [**T**]{.fragment .fade-in fragment-index="1"} Classifying emails as
    spam or ham
-   [**E**]{.fragment .fade-in fragment-index="1"} Watching you label
    emails as spam or ham
    
```{r}
#| echo: false
countdown::countdown(minutes  = 1L)
```

## Statistical Learning vs Machine Learning vs Data Science

-   Machine learning arose as a sub-field of Artificial Intelligence
    which is a sub-fields of Computer Science
-   Statistical learning arose as a sub-field of Statistics
-   There is much overlap, a great deal of "cross-fertilization"
-   "Data Science" - Reflects the fact that both statistical and machine
    learning are about data
-   "Machine Learning" or "Data Science" are "fancier" terms

## Statistics vs Machine Learning

-   Statistics: more concerned with answering why and how things work,
    making inferences
-   Machine/Statistical learning: more concerned with making predictions

## Terminology/Notation {.smaller}

**Ames Housing dataset** - Contains data on 881 houses in Ames, IA. We
are interested in predicting sale price.

The first ten **observations** are shown below.

```{r}
#| echo: false

ames <- readRDS("../data/AmesHousing.rds")
# ames <- ames %>% select(c(1,2,3,5,8,9))
head(ames, 10) |>
  select(1:9) |> 
  kable("html") |> 
  kable_styling(font_size = 12)
```

## Terminology/Notation {.smaller}

**Default dataset** - Contains credit card default data on 10,000
individuals. We are interested in predicting whether somebody will
default or not.

Ten **observations** are shown below.

```{r}
#| echo: false
set.seed(1)
data("Default")
Default |> 
  slice_sample(n=10) |> 
  kable("html") |> 
  kable_styling(font_size=12)
```

## Terminology/Notation

-   **Response/Target/Outcome** - variable we are interested in
    predicting, denoted as $Y$
-   **Features/Inputs/Predictors** - variables used to predict the
    response, denoted as $X$
-   **Feature Matrix** - all features taken together, denoted as
    $\mathbf{X}$
-   Number of data points/observations denoted as $n$
-   Number of features/inputs/predictors denotes as $p$
-   Missing entries in R are denoted as `NA`

## [Question!!!]{style="color:blue"} {.smaller}

For the **Ames Housing** and **Default** datasets:

::: panel-tabset
## Questions

-   What are the corresponding values of $n$ and $p$?
-   What will be the dimension of the corresponding response vector $Y$?
-   What is the value of the 3rd feature for the 2nd observation?

## Answers

-   What are the corresponding values of $n$ and $p$?
    -   Ames: $n = 881$ and $p = 9$
    -   Default: $n = 10000$ and $p = 4$
-   What will be the dimension of the corresponding response vector $Y$?
    -   Ames: $881\times 1$
    -   Default: $10000\times 1$
-   What is the value of the 3rd feature for the 2nd observation?
    -   Ames: `Attchd`
    -   Default: `397.5425`
:::

## [Question!!!]{style="color:blue"} {.smaller}

Suppose you have information about 867 cancer patients on their [age,
tumor size, clump thickness of the tumor, uniformity of cell size, and
whether the tumor is malignant or benign]{.fragment .highlight-red
fragment-index="2"}. Based on these data, you are interested in building
a model to predict the type of tumor (malignant or benign) for future
cancer patients.

-   What are the values of $n$ and $p$ in this dataset?
    [$n = 867, p = 5$]{.fragment .fade-in fragment-index="1"}
-   What are the inputs/features?

## Supervised vs Unsupervised Learning

![Machine Learning Tasks (from Bunker and Fayez,
2017)](images/01/Supervised-Learning-versus-Unsupervised-Learning-Mathworks-nd.png){.r-stretch}

## Supervised Learning

-   We have access to **labeled** data
-   Objective: learn overall pattern of relationship between the inputs
    ($\mathbf{X}$) and response ($Y$) in order to
    -   Investigate the relationship between inputs and response
    -   Predict for potential unseen **test** cases
    -   Assess the quality of predictions

## Types of Supervised Learning

Supervised Learning problems can be categorized into:

::: incremental
-   **Regression** problems (response is quantitative, continuous)
-   **Classification** problems (response is qualitative, categorical)
-   What if I have **discrete** quantitative data (e.g. counts)?
    +   Can use either but one is probably better...
    +   How many discrete values are there?
        +   Enough we can think of the response as continuous? 
    +   What is our goal?
        +   How important is being *exactly* correct?
:::

## Unsupervised Learning

-   No response/outcome variable, just $\mathbf{X}$
-   Understand structure within data
    -   find similar groups of observations based on features
        (**clustering**)
    -   find a smaller subset of features with the most variation
        (**dimensionality reduction**)
-   No gold-standard
-   Easier to collect unlabeled data
-   Useful pre-processing step for supervised learning

## Unsupervised Learning {.smaller}

**US Arrests dataset** - Data on arrests for 50 US states.

The first ten observations are shown below.

```{r}
#| echo: false

data("USArrests")
head(USArrests, 10) |> 
  kable("html") |> 
  kable_styling(font_size = 16)
```

## [Question!!!]{style="color:blue"}

::: {style="font-size: 60%;"}
For each of the following, identify whether the problem belongs to the
supervised or unsupervised learning paradigm

-   Examine the statistics of two football teams, and predict which team
    will win tomorrow's match (given historical data of teams'
    wins/losses to learn from) [**supervised**]{.fragment .fade-in}
-   Given genetic (DNA) data from a person, predict the probability of
    the person developing diabetes over the next 10 years
    [**supervised**]{.fragment .fade-in}
-   Take a collection of 1000 essays written on the US economy, and find
    a way to automatically group these essays into a small number of
    groups of essays that are somehow "similar" or "related"
    [**unsupervised**]{.fragment .fade-in}
-   Examine data on the income and years of education of adults in a
    neighborhood and build a model to predict the income from years of
    education [**supervised**]{.fragment .fade-in}
:::

## Recap {.smaller}

-   What is a Machine Learning algorithm?
    +   T: has task
    +   P: performance is measured
    +   E: improves with experience
-   Terminology:
    +   Features
        +   $p$
    +   Target
    +   Observations
        + $n$
-   Supervised vs. unsupervised learning
    +   Supervised: data is labeled
    +   Unsupervised: data is unlabeled



