---
title: 'MATH 427: Regularization & Model Tuning'
author: Eric Friedlander
footer: "[🔗 MAT 427 - Spring 2025 -  Schedule](https://mat427sp25.netlify.app/schedule)"
logo: "../images/logo.png"
format: 
  revealjs:
    theme: slides.scss
    multiplex: false
    transition: fade
    slide-number: false
    incremental: false 
    chalkboard: true
execute:
  freeze: auto
  echo: true
  cache: true
knitr:
  opts_chunk: 
    R.options:      
    width: 200
---


## Computational Set-Up

```{r}
library(tidyverse)
library(tidymodels)
library(knitr)
library(fivethirtyeight) # for candy rankings data

tidymodels_prefer()

set.seed(427)
```

## Data: Candy {.smaller}

The data for this lecture comes from the article FiveThirtyEight [*The Ultimate Halloween Candy Power Ranking*](https://fivethirtyeight.com/features/the-ultimate-halloween-candy-power-ranking) by Walt Hickey. To collect data, Hickey and collaborators at FiveThirtyEight set up an experiment people could vote on a series of randomly generated candy match-ups (e.g. Reese's vs. Skittles). Click [here](http://walthickey.com/2017/10/18/whats-the-best-halloween-candy/) to check out some of the match ups.

The data set contains 12 characteristics and win percentage from 85 candies in the experiment.

## Data: Candy

```{r}
glimpse(candy_rankings)
```

## Data Cleaning

```{r}
candy_rankings_clean <- candy_rankings |> 
  select(-competitorname) |> 
  mutate(sugarpercent = sugarpercent*100, # convert proportions into percentages
         pricepercent = pricepercent*100, # convert proportions into percentages
         across(where(is.logical), ~ factor(.x, levels = c("FALSE", "TRUE")))) # convert logicals into factors
```

## Data Cleaning

```{r}
glimpse(candy_rankings_clean)
```

## Data Splitting

```{r}
candy_split <- initial_split(candy_rankings_clean, strata = winpercent)
candy_train <- training(candy_split)
candy_test <- testing(candy_split)
```

# Lasso & Ridge Regression

## Question

-   What criteria do we use to fit a linear regression model? Write down an expression with $\hat{\beta_i}$'s, $x_{ij}$'s, and $y_j$'s in it.

## OLS {.smaller}

-   Ordinary Least Squares Regression:

$$
\begin{aligned}
\hat{\beta} =\underset{\hat{\beta}_0,\ldots, \hat{\beta}_p}{\operatorname{argmin}}SSE(\hat{\beta}) &= \underset{\hat{\beta}_0,\ldots, \hat{\beta}_p}{\operatorname{argmin}}\sum_{j=1}^n(y_j-\hat{y}_j)^2\\
&=\underset{\hat{\beta}_0,\ldots, \hat{\beta}_p}{\operatorname{argmin}}\sum_{j=1}^n(y_j-\hat{\beta}_0-\hat{\beta}_1x_{1j} - \cdots - \hat{\beta}_px_{pj})^2
\end{aligned}
$$

-   $\hat{\beta} = (\hat{\beta}_0,\ldots, \hat{\beta}_p)$ is the vector of all my coefficients
-   $\operatorname{argmin}$ is a function (operator) that returns the *arguments* that minimize the quantity it's being applied to

## Ridge Regression {.smaller}

$$
\begin{aligned}
\hat{\beta} &=\underset{\hat{\beta}_0,\ldots, \hat{\beta}_p}{\operatorname{argmin}} \left(SSE(\hat{\beta}) + \lambda\|\hat{\beta}\|_2^2\right) \\
&= \underset{\hat{\beta}_0,\ldots, \hat{\beta}_p}{\operatorname{argmin}}\left(\sum_{j=1}^n(y_j-\hat{y}_j)^2 + \lambda\sum_{i=1}^p \hat{\beta}_i^2\right)\\
&=\underset{\hat{\beta}_0,\ldots, \hat{\beta}_p}{\operatorname{argmin}}\left(\sum_{j=1}^n(y_j-\hat{\beta}_0-\hat{\beta}_1x_{1j} - \cdots - \hat{\beta}_px_{pj})^2 + \lambda\sum_{i=1}^p \hat{\beta}_i^2\right)
\end{aligned}
$$

-   $\lambda$ is called a "tuning parameter" and is not estimated from the data (more on this later)
-   Idea: penalize large coefficients
-   If we penalize large coefficients, what's going to happen to to our estimated coefficients? [[THEY SHRINK!]{.span style="color:red;"}]{.fragment .fade-in}
-   $\|\cdot\|_2$ is called the *$L_2$-norm*

## But... but... but why?!? {.smaller}

::: incremental
-   Recall the Bias-Variance Trade-Off
-   Our reducible error can partitioned into:
    +   Bias: how much $\hat{f}$ misses $f$ by *on average*
    +   Variance: how much $\hat{f}$ moves around from sample to sample
-   Ridge: increase bias a little bit in exchange for large decrease in variance
-   As we increase $\lambda$ do we increase or decrease the penalty for large coefficients?
-   As we increase $\lambda$ do we increase or decrease the *flexibility* of our model?
:::

## LASSO {.smaller}

$$
\begin{aligned}
\hat{\beta} &=\underset{\hat{\beta}_0,\ldots, \hat{\beta}_p}{\operatorname{argmin}} \left(SSE(\hat{\beta}) + \lambda\|\hat{\beta}\|_1\right) \\
&= \underset{\hat{\beta}_0,\ldots, \hat{\beta}_p}{\operatorname{argmin}}\left(\sum_{j=1}^n(y_j-\hat{y}_j)^2 + \lambda\sum_{i=1}^p |\hat{\beta}_i|\right)\\
&=\underset{\hat{\beta}_0,\ldots, \hat{\beta}_p}{\operatorname{argmin}}\left(\sum_{j=1}^n(y_j-\hat{\beta}_0-\hat{\beta}_1x_{1j} - \cdots - \hat{\beta}_px_{pj})^2 + \lambda\sum_{i=1}^p |\hat{\beta}_i|\right)
\end{aligned}
$$

-   LASSO: **L**east **A**bsolute **S**hrinkage and **S**election **O**perator
-   $\lambda$ is called a "tuning parameter" and is not estimated from the data (more on this later)
-   Idea: penalize large coefficients
-   If we penalize large coefficients, what's going to happen to to our estimated coefficients [THEY SHRINK!]{.span style="color:red;"}
-   $\|\cdot\|_1$ is called the *$L_1$-norm*

## Question

-   What should happen to our coefficients as we *increase* $\lambda$?

## Ridge vs. LASSO

::: incremental
-   Ridge has a *closed-form solution*... how might we calculate it?
-   Ridge has some nice linear algebra properties that makes it EXTREMELY FAST to fit
-   LASSO has no *closed-form solution*... why?
-   LASSO coefficients estimated *numerically*... how?
    +   Gradient descent works (kind of) but something called **coordinate descent** is typically better
-   MOST IMPORTANT PROPERTY OF LASSO: it *induces sparsity* while Ridge does not
:::

## Sparsity in Applied Math {.smaller}

-   **sparse** typically means "most things are zero"
-   Example: sparse matrices are matrices where most entries are zero
    +   for large matrices this can provide HUGE performance gains
-   LASSO *induces sparsity* by setting most of the parameter estimates to zero
    +   this means it fits the model and does feature selection SIMULTANEOUSLY
-   Let's do some board work to see why this is...

## LASSO and Ridge in R

```{r}
ols <- linear_reg() |> 
  set_engine("lm")
ridge_0 <- linear_reg(mixture = 0, penalty = 0) |> # penalty set's our lambda
  set_engine("glmnet")
ridge_1 <- linear_reg(mixture = 0, penalty = 1) |> # penalty set's our lambda
  set_engine("glmnet")
ridge_10 <- linear_reg(mixture = 0, penalty = 10) |> # penalty set's our lambda
  set_engine("glmnet")
lasso_0 <- linear_reg(mixture = 1, penalty = 0) |> # penalty set's our lambda
  set_engine("glmnet")
lasso_1 <- linear_reg(mixture = 1, penalty = 1) |> # penalty set's our lambda
  set_engine("glmnet")
lasso_10 <- linear_reg(mixture = 1, penalty = 10) |> # penalty set's our lambda
  set_engine("glmnet")
```

## Create Recipe

```{r}
lm_preproc <- recipe(winpercent ~ . , data = candy_train) |> 
  step_dummy(all_nominal_predictors()) |> 
  step_normalize(all_numeric_predictors())
```

## Question

-   Why do we need to normalize our data?

## Create workflows

```{r}
generic_wf <- workflow() |> add_recipe(lm_preproc)
ols_wf <- generic_wf |> add_model(ols)
ridge0_wf <- generic_wf |> add_model(ridge_0)
ridge1_wf <- generic_wf |> add_model(ridge_1)
ridge10_wf <- generic_wf |> add_model(ridge_10)
lasso0_wf <- generic_wf |> add_model(lasso_0)
lasso1_wf <- generic_wf |> add_model(lasso_1)
lasso10_wf <- generic_wf |> add_model(lasso_10)
```

## Fit Models

```{r}
ols_fit <- ols_wf |> fit(candy_train)
ridge0_fit <- ridge0_wf |> fit(candy_train)
ridge1_fit <- ridge1_wf |> fit(candy_train)
ridge10_fit <- ridge10_wf |> fit(candy_train)
lasso0_fit <- lasso0_wf |> fit(candy_train)
lasso1_fit <- lasso1_wf |> fit(candy_train)
lasso10_fit <- lasso10_wf |> fit(candy_train)
```

## Collect coefficients

```{r}
all_coefs <- bind_cols(model = "ols", tidy(ols_fit)) |> 
  bind_rows(bind_cols(model = "ridge0", tidy(ridge0_fit))) |> 
  bind_rows(bind_cols(model = "ridge1", tidy(ridge1_fit))) |> 
  bind_rows(bind_cols(model = "ridge10", tidy(ridge10_fit))) |> 
  bind_rows(bind_cols(model = "lasso0", tidy(lasso0_fit))) |> 
  bind_rows(bind_cols(model = "lasso1", tidy(lasso1_fit))) |> 
  bind_rows(bind_cols(model = "lasso10", tidy(lasso10_fit)))
```

## Question

::: incremental
-   What should we expect from `ols`, `ridge0`, and `lasso0`?
-   They should be the same!
:::

## Visualize

```{r}
all_coefs |> 
  filter(model %in% c("ols", "ridge0", "lasso0")) |> 
  ggplot(aes(y = term, x = estimate, fill = model, color = model)) +
  geom_bar(position = "dodge", stat = "identity")
```

## Uh-Oh!

::: incremental
-   They're not the same
-   Any idea why?
-   Algorithm used to fit model
    +   `lm` estimates coefficients *analytically*
    +   `glmnet` estimates coefficients *numerically* using an algorithm named "coordinate-descent"
-   Moral: you must understand theory AND application
:::

## Visualizing Coefficients: Ridge

```{r}
all_coefs |> 
  filter(model %in% c("ols", "ridge1", "ridge10")) |> 
  ggplot(aes(y = term, x = estimate, fill = model, color = model)) +
  geom_bar(position = "dodge", stat = "identity")
```

## Visualizing Coefficients: LASSO

```{r}
all_coefs |> 
  filter(model %in% c("ols", "lasso1", "lasso10")) |> 
  ggplot(aes(y = term, x = estimate, fill = model, color = model)) +
  geom_bar(position = "dodge", stat = "identity")
```

## Ridge Coefficients vs. Penalty ( $\lambda$ )

```{r}
ridge0_fit |> extract_fit_engine() |> autoplot()
```

## LASSO Coefficients vs. Penalty ( $\lambda$ )

```{r}
lasso0_fit |> extract_fit_engine() |> autoplot()
```


## Why should I learn math for data science?

-   Today:
    +   Ordinary Least Squares: Linear algebra, Calc 3
    +   LASSO: Calc 3, Geometry, Numerical Analysis
    +   Ridge: Linear Algebra, Calc 3
    +   Bias-Variance Trade-Off: Probability
    
## Question

-   How do you think we should choose $\lambda$?



