---
title: "Computing access"
---

Not that we will go through this as a call and on your first homework. Feel free to revisit it here if you need to.

## R & RStudio

It is highly recommended that you install R and RStudio on your own personal computer. Follow the directions [here](https://posit.co/download/rstudio-desktop/) to install R and RStudio on your computer

If, for some reason, you are unable to use R or RStudio on your personal computer, you may use the [College of Idaho's RStudio Server](https://rstudio.collegeofidaho.edu/). However, I do not recommend you do this as I want you to practice installing packages in this course and you do not have the permissions to do that on the server. If you must use the RStudio Server, please notify Dr. Friedlander *immediately*.

## Git & Github

You need to create a [GitHub](https://github.com/) account. You may want to use this to show-off work to future employers so I recommend using something professional (like your name) as your user name. Once you have done this, email it to Dr. Friedlander so he can add you to the GitHub Classroom.

### Set up your SSH Key

You will authenticate GitHub using SSH. Below are an outline of the authentication steps.

::: callout-note
You only need to do this authentication process one time on a single system.
:::

-   **Step 0**: Type `install.packages("credentials")` into your console.
-   **Step 1**: Type `credentials::ssh_setup_github()` into your console.
-   **Step 2**: R will ask "No SSH key found. Generate one now?" Click 1 for yes.
-   **Step 3**: You will generate a key. It will begin with "ssh-rsa...." R will then ask "Would you like to open a browser now?" Click 1 for yes.
-   **Step 4**: You may be asked to provide your username and password to log into GitHub. This would be the ones associated with your account that you set up. After entering this information, paste the key in and give it a name. You might name it in a way that indicates where the key will be used, e.g., `mat427`)

You can find more detailed instructions [here](https://github.com/DukeStatSci/github_auth_guide) if you're interested.

### Configure git

There is one more thing we need to do before getting started on the assignment. Specifically, we need to configure your git so that RStudio can communicate with GitHub. This requires two pieces of information: your name and email address.

To do so, you will use the `use_git_config()` function from the `usethis` package. You will need to install the `usethis` package in the same way you installed the `credentials` packages above.

Type the following lines of code in the **console** in RStudio filling in your name and the email address associated with your GitHub account.

```{r}
#| eval: false
usethis::use_git_config(
  user.name = "Your name", 
  user.email = "Email associated with your GitHub account")
```

For example, mine would be

```{r eval=FALSE}
usethis::use_git_config(
  user.name = "EricFriedlander",
  user.email = "efriedlander@collegeofidaho.edu")
```

You are now ready interact between GitHub and RStudio!
